Synergistic effects of deleting the tyrosine phosphatases Shp1 and Shp2 on megakaryopoiesis and thrombopoiesis in mice

Version 1

Abstract

The Src homology 2 (SH2) domain-containing non-transmembrane protein-tyrosine phosphatases 1 and 2 (Shp1 and Shp2) have been implicated in regulating signaling from a variety of receptors and cell types, including the thrombopoietin (Tpo) receptor Mpl in megakaryocytes (MKs) and platelets. We previously showed that deletion of Shp1 and Shp2 in the MK/platelet lineage in mice using the Pf4-Cre transgene/loxP system impairs megakaryopoiesis and thrombopoiesis. However, we also observed unexpected phenotypes including a motheaten-like phenotype in Shp1-deficient mice and severe myelofibrosis in mice lacking both phosphatases. To determine whether these were lineage-specific effects, we utilized the Gp1ba-Cre transgenic mouse to delete loxP-flanked Shp1 and Shp2 in mice. Bone marrow-derived MKs from these mice expressed approximately 20-25% of Shp1 and Shp2, whereas platelets contain 5-10% of each phosphatase compared with controls. Minor MK/platelet defects were observed in mice lacking either Shp1 or Shp2 alone, however mice lacking both Shp1 and Shp2 exhibited macrothrombocytopenia, mild bleeding following tail injury, and impaired GPVI-mediated platelet aggregation and Syk phosphorylation, associated with reduction GPVI and integrin α2 subunit expression. Reduced Shp1 and Shp2 expression resulting in a significant reduction in ploidy, a block in MK maturation and proplatelet-producing MKs. Tpo-mediated Ras/MAPK signaling was reduced in Shp1/2-deficient MKs. Treatment of MKs with structurally distinct Shp2 allosteric inhibitors recapitulated key aspects of the Shp2-deficient phenotype, including aberrant megakaryopoiesis and reduced Mpl signaling. Our study highlights the synergistic functions of Shp1 and Shp2 in the MK/platelet lineage, and identifies Shp2 as a potential therapeutic target in myeloproliferative neoplasms.

Key Points

Deletion of Shp1 and Shp2 in the MK/platelet lineage in mice results in macrothrombocytopenia and minor effects on platelet function.

Defects can be partially explained by reduced Mpl signaling and aberrant megakaryopoiesis in the absence of Shp2 activity.

Introduction

Platelets, or thrombocytes are small, anucleate blood cells derived from megakaryocytes (MKs) that are essential for hemostasis and thrombosis.1 MK development, or megakaryopoiesis and platelet production, or thrombopoiesis are tightly regulated processes that take place in the bone marrow (BM), spleen and lungs, but the molecular mechanisms regulating these inter-related processes remain incompletely defined.2,3 The primary driver of megakaryopoiesis, which encompasses differentiation of hematopoietic stem and progenitor cells (HSPCs) to MKs, proliferation, endomitosis and maturation is the cytokine thrombopoietin (Tpo) produced by the kidneys, liver and BM acting on its receptor, myeloproliferative leukemia protein (Mpl) expressed throughout the MK/platelet lineage.4 Tpo does not however explain which MKs undergo proplatelet formation and thrombopoiesis, which occurs when MKs come in contact with sinusoidal blood vessels. Several other cytokines and growth factors, extracellular matrix (ECM) proteins and chemokines are also required for optimal megakaryopoiesis and thrombopoiesis, as evidenced in mice lacking either Tpo or Mpl, which still produce MKs and platelets, albeit at much lower levels than normal mice.5,6

The three primary signaling pathways downstream of Mpl are Janus kinase 2 (JAK2)/Signal transducer and activator of transcription (STAT), Ras/mitogen-activated protein kinase (MAPK) and phosphatidylinositol 3-kinase (PI3K)/AKT, all of which culminate in nuclear translocation of signaling/transcription factors and gene regulation.7 Cytoskeletal remodeling via the RhoA/Rho-associated protein kinases (ROCK) pathway is also an integral part of megakaryopoiesis and thrombopoiesis, notably during endomitosis8, but the molecular link with the Mpl receptor is undefined. Clonal gain-of-function mutations in the Mpl/JAK2/STAT pathway in HSPCs are the most common causes of myeloproliferative neoplasms (MPN)9, characterized by increase MK/platelet production (essential thrombocythemia, ET), increased red blood cell production (polycythemia vera, PV) and associated fibrotic deposits (primary myelofibrosis, PMF).10–12 Notably, the valine 617 phenylalanine (V617F) mutation in the pseudo-kinase domain of JAK2, resulting in increased catalytic active is the most prevalent, found in 60% of MPN cases, followed by C-terminal deletions in calreticulin, resulting in the release of truncated calreticulin and it acting as a ligand of Mpl in 30% of the cases, and mutations in Mpl resulting in increased expression in <5% of the cases.13 JAK2 and MAPK kinase (MEK) inhibitors are effective in some instances of MPN14, however there is a tendency of resistance and relapse, requiring alternative and improved therapeutics.

The Src homology 2 (SH2) domain-containing non-transmembrane protein-tyrosine phosphatases (PTPs) 1 and 2 (Shp1 and Shp2) are involved in hematopoietic cell differentiation, proliferation, function and survival, including megakaryopoiesis and thrombopoiesis.15 Intra- and inter-molecular interactions of the tandem SH2 domains with the catalytic domain and phospho-tyrosine binding partners regulate the activity and compartmentalization of both phosphatases, respectively. Tandem phospho-tyrosine residues in the C-terminal tails of Shp1 and Shp2 are implicated in substrate binding and regulation.16 Despite high structural similarities, the two phosphatases have distinct biological functions. Shp1, encoded by the PTPN6 gene, is primarily expressed in hematopoietic cells and is a negative regulator of immune receptor signaling, particularly the immunoreceptor tyrosine-based activation motif (ITAM)/Spleen tyrosine kinase (Syk)/phospholipase pathway.17 Shp2, encoded by the PTPN11 gene, is ubiquitously expressed and is a positive regulator of cytokine and growth factor signaling, specifically the Ras/MAPK and PI3K/AKT pathways.18 However, their functions go beyond these pathways and are context dependent. Downstream substrates and effectors of both phosphatases remain elusive.

We previously showed that targeted deletion of Shp1 and Shp2 in the MK/platelet lineage using the Pf4-Cre transgene/loxP system impairs megakaryopoiesis and thrombopoiesis in mice.15 However, we also found that these mice exhibited unexpected, non-MK/platelet-related defects, including a motheaten-like phenotype in Shp1-deficient mice, and severe myelofibrosis in mice lacking both phosphatases. Intriguingly, the motheaten-like phenotype was lost in Shp1/2 double-deficient mice, suggesting that Shp1 and Shp2 have opposing effect in this pathology. To determine whether these were in fact MK/platelet lineage-specific effects, we utilized the Gp1ba-Cre transgenic mouse to delete loxP-flanked Shp1 and Shp2 in mice. In contrast to Pf4-Cre transgenic mice, conditional deletion of Shp1 or Shp2 using the Gp1ba-Cre transgene had little if any detectable effect on megakaryopoiesis or thrombopoiesis in these mice under steady state conditions. However, deletion of both Shp1 and Shp2 using Gp1ba-Cre resulted in macrothrombocytopenia and aberrant MK development and maturation, due to a block in Mpl-mediated Ras/MAPK signaling. GPVI-induced platelet aggregation was also impaired due to reduced GPVI and α2 integrin subunit expression. Similar effects were seen with the structurally-distinct allosteric Shp2 inhibitors, SHP099 and RMC-4550.19 Findings demonstrate synergistic effects of deleting Shp1 and Shp2 in the MK/platelet lineage in mice by disrupting distinct signaling pathways, and highlight Shp2 as a potential therapeutic target in MPN.20,21

Materials and Methods

Mouse models

All mice used were on a C57BL/6 background. Ptpn6fl/fl, Ptpn11fl/fl and Gp1ba-Cre+/KI mice were generated, as previously described.22–24 MK/Platelet specific Shp1 and Shp2 knockout (KO) mice were generated by crossing Ptpn6fl/fl and Ptpn11fl/fl mice with the Gp1ba-Cre transgenic deleter mouse Gp1ba-CreKI/+ mice. Gp1ba-Cre KI/+ were used as control mice (CT) (Supplemental Figure S1). All animal experiments were conducted in accordance with the CREMEAS Committee on the Ethics of Animal Experiments of the University of Strasbourg (Permit Number: E67-482-10, Project approval number: APAFIS#28221-2020111714459066).

Inhibitors

SHP099 and RMC-4550 Shp2 allosteric inhibitors were purchased from MedChemExpress. M029 is a covalent allosteric Shp1 inhibitor (Z.Q. and Z.Y.Z, manuscript in preparation) and F2Ac is a reversible active site directed Shp1 inhibitor (J.M. and Z.Y.Z., manuscript in preparation.

Platelet preparation and functional assays

Blood was collected from the aorta of anesthetized mice into 1/10 acid-citrate-dextrose (ACD) anticoagulant. Platelet counts were measured from peripheral blood samples using an automated hematology analyzer (Element HT5 counter, Heska company). Washed platelets were prepared as previously described.25 Platelet counts were normalized and used for aggregation (2 x 108/ml) or biochemical analysis (5 x 108/ml). Platelet aggregation was measured using the lumi-aggregometer APACT®4004. Surface glycoprotein expression was measured in whole blood by flow cytometry using fluorescein isothiocyanate (FITC)-conjugated antibodies. Resting and activated platelets were fixed and stained with anti-P-selectin antibody.

Megakaryocyte culture and functional assays

Mature MKs from mouse BM were cultured and analysed, as previously described.26 BM cells were obtained from femora, tibiae and iliac crests of C57BL6 mice for pharmacological study and Gp1ba-Cre mice for other experiments by flushing, and Lin-cells obtained after immunomagnetic sorting were cultured for 3 days at 37°C 5% CO2 at a concentration of 1×106 cells/well in a Dulbecco’s Modified Eagle Medium (DMEM), 10% Fetal Bovine Serum (FBS), 50 ng/ml Tpo, 100 U/ml Hirudin, 1% PSG.26 MK ploidy was assessed after 3 days of culture, using FITC-coupled anti-CD41 antibody and DNA staining with propidium iodide (PI). Samples were acquired using the BD Fortessa X20 and analyzed using BD FACSDiva software. The ability of MKs to form proplatelets was assessed in vitro, and ex vivo using the explant assay.27

Immunoblotting

Platelet and MK whole cell lysates (WCLs) were prepared and analyzed by automated capillary-based immunoassay (ProteinSimple Jess), prepared according to manufacturer’s instructions, as previously described.28 The High Dynamic Range profile was used for chemiluminescent and fluorescent multiplexing signal detection. Optimized antibody dilutions and sample concentrations used are provided in Supplemental Table 1.

Immunohistochemistry

Spleens from KO and litter-matched CT mice were fixed in buffered formalin and embedded in paraffin. Sections (5 μm) were H&E stained and examined by light microscopy using a 40× objective.

Electron microscopy

Maturation stage of MKs was assessed by electron microscopy (EM). BM samples obtained by flushing mouse femora with 0.1 M sodium cacodylate buffer were fixed in 2.5% glutaraldehyde and embedded in Epon as previously described.29 Thin sections were stained with uranyl acetate and lead citrate, and examined under a JEOL 2100Plus transmission electron microscope at 120 kV (Jeol, Tokyo, Japan). MKs at stages I, II and III were identified using distinct ultrastructural characteristics: stage I, a cell 10–15 μm in diameter with a large nucleus; stage II, a cell 15–30 μm in diameter containing platelet-specific granules; stage III, 40 µm, a MK containing a well-developed demarcation membrane system defining cytoplasmic territories and a peripheral zone. Samples from at least three mice of each genotype were examined in each case.

Platelet recovery assay

An intraperitoneal injection of anti-GPIbα antibody (Emfret) at 2 mg/kg is performed at D0. Blood samples were collected and platelet counts measured prior to injection, and at D5, D10, D15 and D20.30

HSPC colony-forming unit (CFU) assay

Cells are obtained by flushing and were resuspended in IMDM and plated at a density of 1×10⁴ cells per dish in methylcellulose-based medium (MethoCult™ GF M3434, Stemcell Technologies) containing appropriate cytokines to support multilineage hematopoietic colony formation. Cultures were incubated at 37 °C with 5% CO₂ in a humidified incubator for 10–14 days. Colonies were scored under an inverted microscope and classified based on morphological criteria into CFU-G (granulocyte), CFU-M (macrophage), CFU-GM (granulocyte-macrophage), BFU-E (erythroid), and CFU-GEMM (granulocyte, erythrocyte, macrophage, megakaryocyte). Only colonies containing ≥50 cells were counted. Results were expressed as the mean colony number ± SEM from n = 3 independent experiments.

Statistical analysis

Statistical analysis was performed using GraphPad Prism version 9 software. Data were expressed as mean ± standard error mean (SEM). Statistical significance was analyzed by one- or two-way ANOVA followed by the appropriate post hoc test, or Kruskal-Wallis test for nonparametric data, a Bonferroni correction was applied for multiple comparisons, as indicated in figure legends. P-values < 0.05 were considered statistically significant.

Results

Deletion of Shp1 and Shp2 in the MK/platelet lineage

To validate the efficiency of protein ablation of Shp1 and Shp2 proteins in MKs/platelets using the Gp1ba-Cre transgenic mouse strain, we employed a quantitative capillary-based immunoassay. This was performed on cells derived from Ptpn6fl/fl;Gp1ba-Cre+/KI (Shp1 KO), Ptpn11fl/fl;Gp1ba-Cre+/KI (Shp2 KO) and Ptpn6fl/fl;Ptpn11fl/fl;Gp1ba-Cre+/KIdouble-knockout (DKO) mice, with Gp1ba-Cre+/KI littermates serving as controls.

We firstly evaluated Shp1 and Shp2 levels in MK-sorted progenitors (MKPs) in order to obtain a pure population of matures MKs. Under these conditions, we quantified protein loss of Shp1 and Shp2 in MKs, as shown in Figure 1Ai-iii, with significant 70% and 82% reduction of Shp1 protein in Shp1 KO and Shp1/2 DKO mice, respectively and 78% and 75% reduction of Shp2 protein in Shp2 KO and Shp1/2 DKO mice, respectively compared to controls.

Protein expression of Shp1 and Shp2 in megakaryocytes and platelets from Gp1ba-Cre mice.

(i) Representative blots of Shp1 and Shp2 expression levels from CT, Shp1, Shp2 and Shp1/2 DKO mice by capillary immunoassays with the respective antibodies in (A) sorted-MK progenitors and (B) washed platelets. Percentage of residual level of (ii) Shp1 and (iii) Shp2 in Shp1 KO (blue), Shp2 KO (green) and Shp1/2 DKO (red) mice. n = 3-6 mice per genotype. Mean ± SEM; one way-ANOVA; ** P < 0.01, *** P < 0.001.

We next evaluated protein levels in washed platelets and found efficient deletion in Shp1 KO and Shp2 KO platelets, with a significant 90% and 92% reduction of Shp1 protein in Shp1 KO and Shp1/2 DKO mice respectively, and 95% and 96% reduction of Shp2 protein in Shp2 KO and Shp1/2 DKO mice respectively, compared to controls (Figure 1Bi-iii). In addition, there was no upregulation of Shp1 in Shp2-deficient MKs/platelets and vice versa. These results indicate that the Gp1ba-Cre transgene efficiently deleted loxP-flanked Ptpn6 and Ptpn11, resulting in the significant loss of Shp1 and Shp2 in MKs/platelets.

Macrothrombocytopenia and increased bleeding in Shp1/2 DKO mice

Deletion of Shp1 or Shp2 using the Gp1ba-Cre-mouse strain did not affect platelet counts and volumes as shown in Figure 2Ai-ii and Supplemental Table 2. However, Shp1/2 DKO mice were macrothrombocytopenic, with a 40% reduction in platelet count and 15% increase in platelet volume (Figure 2Ai-ii). Monocyte, neutrophil and eosinophil counts were marginally increased in Shp1/2 DKO mice, suggesting possible immune responses. (Supplemental Table 2).

Macrothrombocytopenia and mild increased bleeding in Shp1/2 DKO mice.

(A) Platelet counts (i) and platelet volumes (ii) of control (CT) (n=75), Shp1 KO (n=21), Shp2 KO (n=29) and Shp1/2 DKO (n=46) mice. Mean ± SEM; one way-ANOVA; *** P < 0.001. (B) (i) Cumulative bleeding time was recorded (ii) and the volume of blood loss was measured over 30 minutes. Mean ± SEM; one way-ANOVA; * P < 0.05. (C) Mean fluorescence intensity (MFI) of platelet (i) GPVI and (ii) α2 integrin expression in CT (n=10), Shp1 KO (n=4), Shp2 KO (n=4) and Shp1/2 DKO (n=13) mice. Mean ± SEM; one way-ANOVA; *** P < 0.001.

The primary function of platelets is to prevent blood loss following injury. Therefore, we investigated the effect of the loss of Shp1 and Shp2 on hemostasis. Shp1/2 DKO exhibited a significant increase in bleeding following tail injury as shown in Figure 2Bi-ii. Bleeding time was mildly prolonged compared to control mice, suggesting impaired hemostasis. Quantitative measurement of blood loss revealed a substantial increase in hemoglobin levels (Figure 2Bi-ii).

Aberrant functional responses of Shp1/2-deficient platelets

We next investigated platelet receptor expression by flow cytometry. Receptor expression was normal in Shp1-and Shp2-deficient platelets. However, we found that the collagen activation receptor GPVI and the α2 subunit of the collagen integrin α2β1 were reduced by 25% and 13%, respectively, in Shp1/2-deficient platelets (Figure 2Ci-ii and Supplemental Table 3). All other major surface receptors analyzed were unaltered in these platelets (Supplemental Figure S2 and Supplementary Table 3).

We next checked platelet aggregation by light transmission aggregometry. Shp1- and Shp2-deficient platelets responded normally to all agonists tested (Figure 3A-C). However, consistent with reduced GPVI and α2 surface expression, Shp1/2-deficient platelets exhibited a marginal reduction in aggregation to a low and intermediate concentration of the GPVI-specific agonist CRP (1 and 3 μg/mL), and collagen (3 μg/mL), which signals via GPVI and binds with high affinity to the integrin α2β1 (Figure 3Ai-ii). However, they responded normally to thrombin (0.06 and 0.1 U/ml) (Figure 3Bi-ii), antibody-mediated cross-linking of the hem-ITAM-containing podoplanin receptor CLEC-2 (3 μg/mL) and ADP (3 μM) (Figure 3Ci-ii).

Aberrant platelet function in Shp1/2 DKO mice.

(A-C) Aggregation of washed platelets were measured by lumi-aggregometry in response to agonists indicated. Representative traces, n= 4-8 mice/genotype per condition, percentage of aggregation at 5 minutes and area under the curve (AUC) quantification. Mean ± SEM, n = 5-17 mice/genotype per condition, one way-ANOVA, * P < 0.05. (D) Mean fluorescence intensity (MFI) of P-selectin expression of control (CT), Shp1 KO, Shp2 KO and Shp1/2 DKO platelets in whole blood in response to (i) 1 and 3 μg/mL CRP and (ii) 100 and 500 μM PAR4 peptide (PAR4p). Mean ± SEM, n = 5-9 mice/genotype per condition, two way-ANOVA; * P < 0.05, *** P < 0.001.

We also measured P-selectin surface expression in whole blood, as a marker of α-granule secretion. Consistent with reduced GPVI expression, Shp1/2-deficient platelets did not respond to 1, 3 and 10 μg/mL CRP (Figure 3Di), but aggregated normally to 100 and 500 μM PAR-4 peptide (Figure 3Dii).

Aberrant ITAM signaling in Shp1- and Shp2-deficient platelets

Since Shp1 and Shp2 have been implicated in regulating ITAM-containing receptor signaling, we investigated GPVI and CLEC-2 signaling in Shp1- and Shp2-deficient platelets. We focused on SFKs, which phosphorylate tyrosine residues within the ITAM of GPVI-associated FcR γ-chain and hemi-ITAM of the CLEC-2 receptor, and Syk, which docks to both via its tandem SH2 domains, and mediates downstream effects. SFK activation was indirectly measured as trans-autophosphorylation of the highly conserved tyrosine residue 418 in Src (Src p-Tyr418) and Syk activation was measured as trans-autophosphorylation of tyrosine residues 519/520 (p-Tyr519/20), which directly correlate with activity. Western blot analysis demonstrated that Syk p-Tyr519/520 was significantly decreased in CRP–stimulated platelets from Shp1/2-deficient platelets (Figure 4Ai), and marginally reduced in CLEC-2-stimulated Shp1 KO and Shp2 KO deficient platelets. (Figure 4Bi). Notably, Src p-Tyr418 was not altered under any conditions (Figure 4Aii and Bii), suggesting that Shp1 and Shp2 modulate Syk-dependent signaling downstream of GPVI and CLEC-2.

Aberrant platelet signaling in Shp1/2 DKO mice.

Whole cell lysates of resting and (A) 10 µg/ml CRP-stimulated and (B) 10 µg/ml activating CLEC-2 antibody platelets from Shp1 KO, Shp2 KO and Shp1/2 DKO mice and litter-matched CT mice were western blotted with anti-Src p-Ty418, Src total, -Syk p-Tyr519/520 and Syk antibodies. (i) Representative blots of capillary-based immunoassays and (ii) quantification of peak areas from three independent experiments, Mean ± SEM, n = 3-4 mice/genotype; one-way ANOVA, *** P < 0.001.

Impaired megakaryocyte differentiation and function

To determine the cause of thrombocytopenia in Shp1/2 DKO mice, we measured platelet recovery following antibody-mediated depletion as an indicator of platelet turnover. As expected, administration of the anti-GPIbα antibody resulted in extreme and sustained thrombocytopenia, confirming efficient depletion of circulating platelets (Supplemental Figure S3). No differences were observed the different phases or kinetics of platelet clearance or recovery in any of the mouse models. Interestingly, spleen size and weight were significantly increased in Shp1/2 DKO mice (Supplemental Figure S4), suggesting compensatory extramedullary hematopoiesis. However, HSPC colony-forming unit (CFU) assays performed on BM cells showed no significant difference in the frequency or composition of HSPCs, including MKPs between genotypes in vitro (Supplemental Figure S5).

Despite the maintenance of normal platelet turnover in vivo, in vitro analyses revealed pronounced defects in MK development, maturation and function of Shp1/2-deficient MKs (Figure 5A-C). BM-derived MKs from DKO mice exhibited a severe defect in polyploidization, as shown by a significant increase in 2-8N immature MKs, and decreased 16-64N mature MKs (Figure 5Ai-ii). Moreover, proplatelet formation was severely impaired, with a significantly lower percentage of MKs extending cytoplasmic projections compared to controls (Figure 5B).

Defects in megakaryocyte maturation and Tpo signaling in Shp1/2 DKO.

(A) Mature BM-derived MKs from Shp1 KO, Shp2 KO, Shp1/2 DKO, and litter-matched CT mice were stained with propidium iodide and ploidy of cells was quantified by flow cytometry. (i-ii) The percentage of 2-4N and 8-128N ploidy cells was quantified (n = 4-6 mice/genotype; mean ± SEM; two-way ANOVA, * P < 0.05, ** P < 0.01, *** P < 0.001). (B) Ex vivo proplatelet formation. Percentage of MKs forming proplatelet was quantified in culture. Mean ± SEM, two-way ANOVA, *** P < 0.001. (C) Mature BM-derived MKs from Shp1 KO, Shp2 KO, Shp1/2 DKO and litter-matched CT mice were stimulated with 50 ng/mL thrombopoietin (Tpo) for 10 min at 37°C. Whole cell lysates were western blotted with indicated antibodies. Representative blots capillary-based immunoassays and quantification of peak areas from n = 3 independent experiments/genotype; two-way ANOVA, *** P < 0.001.

Collectively, these data suggest that while homeostatic and compensatory thrombopoiesis is preserved in vivo, potentially supported by extramedullary hematopoiesis, Shp1 and Shp2 are required for efficient megakaryopoiesis in vitro, highlighting a critical role in late-stage megakaryopoiesis under defined conditions.

Reduced Tpo signaling in Shp1/2-deficient MKs

To understand the molecular basis of impaired megakaryopoiesis in Shp1/2 DKO mice, we assessed Tpo-mediated signaling in BM-derived MKs. We previously demonstrated that Shp2 is a positive regulator of ERK1/2 downstream of Mpl. Tpo-mediated ERK1/2 activation was normal in Shp1-deficient MK and significantly reduced in Shp2- and Shp1/2-deficient MKs. However, other downstream effectors, including AKT and STAT3, showed modest or variable changes (Figure 5C). These findings provide a mechanistic explanation for the impaired MK development and maturation observed in vitro, and defective Tpo signaling through the ERK1/2 pathway in Shp1/2-deficient MK, underscoring the importance of these phosphatases in late-stage megakaryopoiesis.

Defective proplatelet formation and MK maturation in Shp1/2-deficient mice

To further characterize the functional consequences of Shp1 and Shp2 deficiency on megakaryopoiesis, we evaluated proplatelet formation ex vivo and examined MK morphology in situ. We first performed ex vivo proplatelet formation assays using BM explants. The results revealed a marked reduction in proplatelet formation in Shp1/2 DKO mice compared with controls, with an increased number of round MKs and a decreased number of MK-forming proplatelets in Shp1/2 DKO mice (Figure 6A-B). To assess MK maturation in situ, we performed EM and ploidy analysis in BM. In control mice, MK displayed characteristic features of terminal maturation, including large polyploid nuclei, demarcation membrane system (DMS) organization, and extensive cytoplasmic development. Shp1/2 DKO MK exhibited a maturation arrest, with increased frequency of stage III MK, defined by incomplete DMS formation, and limited cytoplasmic granularity (Figure 6Ci-ii). This was further supported by flow cytometric analysis of DNA content, showing lower ploidy level (4N–8N) in Shp1/2 DKO MK (Figure 6D). These findings provide further evidence of the involvement of Shp1 and Shp2 in late-stage MK maturation and proplatelet formation, absence of which results in the accumulation of immature, stage III MK in the BM and a functional failure to complete thrombopoiesis.

Defects in ex vivo proplatelet formation and in vivo MK maturation in Shp1/2 DKO mice.

Bone marrow explants. Proportion of MKs from Shp1 KO, Shp2 KO, Shp1/2 DKO and litter-matched CT mice extending proplatelets at 6h of observation were observed. Bars represent the mean ± SEM of six independent experiments. (A) Quantification and (B) representatives’ images of round MKs, MKs with large ends and proplatelet (PPT) forming MKs. Scale bar: 50 µm. Mean ± SEM, one-way ANOVA, * P < 0.1, *** P < 0.001. (C) Classification of the MK according to their maturation stage: stage I (absence of granules), stage II (granules and developing demarcation membrane system (DMS) not yet organized), stage III (DMS organized in cytoplasmic territories). Data are reported as the percentage of the total number of MK, (i) of all stage MK and (ii) only stage 3 MK. Bars represent the mean ± SEM in three BM samples, *** P < 0.001. (D) Ploidy of in situ BM MKs measured by flow cytometry. Mean ± SEM, two-way ANOVA, * P < 0.1, *** P < 0.001.

Shp2 allosteric inhibitors impair megakaryopoiesis and Tpo signaling

To further delineate the respective roles of Shp1 and Shp2 in MK and Tpo signaling, we employed two pharmacologically distinct inhibitors for each phosphatase to ensure specificity and validate findings from genetically modified mouse models.

For Shp1, we used M029, an covalent allosteric inhibitor that induces conformational changes that block Shp1 activity (Z.Q. and Z.Y.Z., manuscript in preparation), and F2Ac, a reversible inhibitor, which directly target the catalytic site (J.M. and Z.Y.Z., manuscript in preparation). Treatment of murine BM-derived MK with these Shp1 inhibitors showed no significant effects on cell viability, proliferation and MK ploidy (Figure 7Ai-iii), nor on Tpo-mediated signaling (Figure 7B), correlating with what was observed in Shp1-deficient MKs. Similarly, Shp2 was inhibited using SHP099 and RMC-4550, both selective allosteric inhibitors that stabilize Shp2 in its inactive conformation31. Addition of Shp2 inhibitors resulted in a marginal decrease in cell viability and a marked reduction of cell proliferation, both established functions of Shp2 in other lineages (Figure 7Ci-ii). Moreover, treatment with either SHP099 or RMC-4550 resulted in a significant inhibition of ploidy with a reduction in the percentage of cells achieving >8N (Figure 7Ciii), and a significant reduction in proplatelet formation (Figure 7Civ). This was accompanied by a severe impairment in Tpo-mediated ERK1/2 and AKT phosphorylation (Figure 7D). This dual-inhibitor profiling reveals that Shp2, but not Shp1, is critical for Tpo-driven MK proliferation, survival, endomitosis and Mpl signaling, highlighting Shp2 as a key regulator of late stage megakaryopoiesis and suggest that its inhibition disrupts thrombopoietic pathways.

Inhibition of Shp2 activity impairs megakaryopoiesis and Mpl signaling.

(A) Effects of two selective Shp1 allosteric inhibitors 10 µM M029 and 10 µM F2Ac were tested on megakaryopoeisis. (i) Viability, (ii) proliferation, and (iii) MK ploidy. Quantification from n = 3-4 independent experiments/condition. (B) Whole cell lysates of resting and 50 ng/mL Tpo-stimulated MKs were western blotted with the indicated antibodies. Representative blots capillary-based immunoassays and quantification of peak areas from n = 3 independent experiments/genotype. (C) Effects of two selective Shp2 allosteric inhibitors 10 µM SHP099 and 10 µM RMC-4550 were tested on megakaryopoeisis. (i) Viability, (ii) proliferation, (iii) MK ploidy and (iv) percentage of MKs forming proplatelets. Quantification from n = 3-6 independent experiments/condition. * P < 0.05; ** P < 0.01; *** P < 0.001, two-way ANOVA. (D) Whole cell lysates of resting and 50 ng/mL Tpo-stimulated MKs were western blotted with the indicated antibodies. Representative blots capillary-based immunoassays and quantification of peak areas from n = 3 independent experiments/genotype. * P < 0.05; ** P < 0.01; two-way ANOVA.

Discussion

In this study, we used the Gp1ba-Cre transgenic mouse model to achieve conditional deletion of the tyrosine phosphatases Shp1 and Shp2 in the MK/platelet lineage. Our results demonstrate efficient deletion of Shp1 and Shp2 in MKs and more completely in platelets, confirming the utility of this mouse model for studying protein involvement in megakaryopoiesis and thrombopoiesis. Residual Shp1 and Shp2 in MKs allowed us to study the phenotype of DKO MKs more thoroughly compared with DKO MKs on the Pf4-Cre background, which were severely compromised developmentally and did not survive in vitro.15 Additionally, the less severe MK defects meant fewer platelet anomalies, which contained lower levels of Shp1 and Shp2 than the MKs, suggesting different half-lives of the phosphatases in MKs versus platelets. The Gp1ba-Cre transgenic mouse allowed us to reveal distinct and synergistic functions of Shp1 and Shp2 in these processes, without confounding effects of deleting Shp1 and Shp2 in non-MK/platelet lineages observed with the Pf4-Cre transgene mouse.

Deletion of Shp1 or Shp2 alone did not significantly affect platelet counts or volumes, while combined deletion resulted in macrothrombocytopenia characterized by a 40% reduction in platelet counts and marginal increase in platelet volume. This phenotype was accompanied by prolonged bleeding and increased blood loss following tail tip excision, indicating impaired hemostatic function. These defects correlated with reduced GPVI and α2 expression, GPVI-mediated platelet aggregation and P-selectin expression. GPVI signaling was also impaired in Shp1/2-deficient platelets, with reduced Syk phosphorylation despite normal SFK activation, suggesting that Shp1 and Shp2 modulate ITAM/Syk proximal signaling. This is further supported by the marginal reduction in Syk activation downstream of the hemi-ITAM-containing CLEC-2 receptor.

Similar phases and kinetics of platelet recovery following immune-induced platelet clearance suggest that thrombocytopenia in Shp1/2 DKO mice is not due to increased platelet clearance, but rather to impaired MK development, maturation and function. In vitro assays showed significant defects in MK polyploidization and proplatelet formation, while in vivo analysis demonstrated an accumulation of immature MKs with incomplete demarcation membrane system formation and lower ploidy levels. These findings point to a critical requirement for Shp1 and Shp2 in late-stage MK maturation, a process tightly regulated by Tpo signaling through Mpl. Consistently, Tpo-mediated activation of the ERK1/2 pathway was markedly reduced in Shp2- and Shp1/2-deficient MKs, while changes in AKT and STAT3 pathways were modest, underscoring the prominent role of Shp2 in facilitating Tpo-driven Ras/MAPK signaling. Based on these findings, the synergistic effects of deleting both Shp1 and Shp2 suggests that Shp1 acts through another, as yet undefined pathway vital for megakaryopoiesis and thrombopoiesis. RhoA/ROCK is an obvious candidate that Shp1 has been implicated in regulating in other lineages and is essential for cytoskeletal remodeling during endomitosis.

Pharmacological inhibition corroborated genetic findings. Selective inhibition of Shp2 with allosteric inhibitors SHP099 and RMC-4550 significantly impaired MK viability, proliferation, polyploidization, proplatelet formation, and Tpo signaling, while Shp1 inhibitors M029 and F2Ac had no significant effects. This demonstrates a dominant role for Shp2 in regulating MK proliferation and differentiation downstream of Tpo, whereas Shp1 is dispensable for these functions. The pharmacological data thus confirm the functional specificity of the two phosphatases and provide further mechanistic insights into their differential contributions. The milder nature of the platelet and hemostasis phenotypes in Shp1/2 DKO platelets suggests that inhibiting Shp2 will not having any bleeding consequence in patients.

The relatively milder, lineage-specific phenotype of Gp1ba-Cre-driven Shp1/2 DKO mice provides distinct advantages to the previously described Pf4-Cre mouse model 32,33 (Supplemental Table 4). Pf4-Cre-mediated deletion induces recombination earlier during megakaryopoiesis, starting in MK progenitors, whereas Gp1ba-Cre is activated predominantly in mature MKs and platelets. This temporal difference allows normal early MK development and platelet production, resulting in less severe thrombocytopenia and functional defects. Additionally, our protein quantification data show significant, but incomplete ablation of Shp1 and Shp2 in BM-derived MKs, suggesting residual phosphatase activity is sufficient under steady-state conditions. In contrast, Pf4-Cre models generally achieve near-complete deletion at earlier stages of development, leading to profound thrombocytopenia and often lethality. Furthermore, compensatory extramedullary hematopoiesis observed in our Gp1ba-Cre model may further alleviate thrombocytopenia. Importantly, this milder phenotype allowed us to study the Shp1/2 DKO mice in detail, which was impossible with the Pf4-Cre model due to the severity of the phenotype and cell viability. Findings further establish the utility of the Gp1ba-Cre strain for dissecting the distinct roles of Shp1 and Shp2 in late-stage megakaryopoiesis and platelet biology.

Collectively, Shp2 is a critical positive regulator of Mpl signaling and downstream Ras/MAPK pathways, essential for MK proliferation, polyploidization, and proplatelet formation, whereas Shp1 plays a more modulatory role in ITAM-containing receptor signaling in platelets without directly impacting MK maturation. The combined deletion of Shp1 and Shp2 reveals additive effects on platelet function and MK development, underscoring the importance of coordinated phosphatase regulation in maintaining platelet homeostasis. This study provides the first detailed analysis of Shp1 and Shp2 functions in megakaryopoiesis and platelet biology using a late-stage MK/platelet-specific deletion model, highlighting the indispensable role of Shp2 in megakaryopoiesis and thrombopoiesis, and distinct contributions of Shp1 in platelet signaling. It also emphasizes the utility of the Gp1ba-Cre mouse for exploring protein function in the MK/platelet lineage. These insights may have broader implications for understanding phosphatase regulation in hematopoiesis and could inform therapeutic targeting of Shp2 in hematologic diseases.

eLife Assessment

This manuscript provides an important contribution to the field of platelet biogenesis, and the convincing evidence will advance our understanding of signal transduction driving the development of late megakaryopoiesis and platelet reactivity that results in bleeding diathesis. The paper is noteworthy for analyzing two related, either singly or in combination, tyrosine phosphatases in this conditional, stage development gene knockout. Because SHP1 is a negative regulator and SHP2 is an activator, the synergistic effects found in the double knockout were surprising.

Reviewer #1 (Public review):

Barré et al. investigated the role of Shp1 and Shp2 in megakaryocytes (MKs) and platelets by conditional knock-out of Shp1, Shp2, or both under the control of the Gp1ba promoter. Deletion of Shp1 and Shp2 in MKs and platelets was almost complete. The Shp1/Shp2 double knock-out mice displayed macrothrombocytopenia and increased bleeding, whereas the single knock-outs did not show significant defects. Platelet function was aberrant in DKOs, but not in single knock-outs, and so was ligand-induced signaling, particularly Syk phosphorylation.

Megakaryocyte maturation was impaired in Shp1/Shp2 DKO mice. Ligand-induced signaling was impaired in Shp2 knock-out and DKO. Ex vivo formation of platelets and in vivo maturation of MKs were impaired in DKO mice. Pharmacological inhibitors of Shp1 and Shp2 had largely similar effects as observed in the single knock-outs. The authors conclude that Shp1 and Shp2 have synergistic functions in the MK/platelet lineage, and that Shp2 may be a potential therapeutic target in myeloproliferative neoplasms.

Strengths:

The data clearly show effects of the Shp1/Shp2 double knock-out on MKs and platelets.

Weaknesses:

There appears to be a discrepancy between the results with the Shp2 single knock-out and the Shp2 inhibitor: the Shp2 knock-out does not affect MKs and platelets, except Erk1/2 signaling, whereas the Shp2 inhibitors appear to affect MK function.

This work is interesting and may have potential from a therapeutic point of view.

Reviewer #2 (Public review):

Summary:

In this manuscript, Barré et al. investigate the roles of the phosphatases Shp1 and Shp2 in the megakaryocyte and platelet lineage using genetic depletion in mice. By employing Gp1ba-Cre-based models, the study builds on the authors' previous work and addresses some limitations associated with earlier Pf4-Cre approaches. The authors report relatively mild alterations in megakaryocyte and platelet parameters in mice lacking either Shp1 or Shp2 alone, whereas combined deletion of both phosphatases results in macrothrombocytopenia, mild bleeding, and impaired GPVI-dependent platelet aggregation accompanied by reduced Syk phosphorylation. The functional platelet defects are linked to reduced expression of GPVI and integrin α2, while thrombocytopenia is associated with impaired megakaryocyte maturation, reduced ploidy, defective proplatelet formation, and altered TPO-dependent Ras/MAPK signaling. Similar effects on megakaryopoiesis are also observed in vitro following treatment with newly developed Shp2 inhibitors.

Strengths and Weaknesses:

The study addresses an important biological question and presents a substantial dataset that could contribute to a better understanding of Shp1 and Shp2 function in platelet biology. However, several aspects of data presentation and interpretation would benefit from additional clarification. In particular, while the authors conclude that single genetic deletion or pharmacological inhibition of Shp1 has a limited impact and that the major phenotypes are specific to combined Shp1/2 deletion or Shp2 inhibition, some of the data suggest more nuanced effects that may warrant further discussion.

Reviewer #3 (Public review):

Summary:

In this manuscript, Barré et al utilize the Gp1ba-Cre transgenic mouse model to build upon previous findings in a Pf4-Cre system to investigate the effects of individual and combined Shp1 and Shp2 deletion in megakaryocytes and platelets. They report decreased megakaryocyte maturation, macrothrombocytopenia, and increased bleeding primarily in association with the Shp1/Shp2 double-knockout condition. The authors further show that this phenotype appears to be driven primarily by Shp2 and implicate dysregulation of Mpl signaling and downstream Ras/MAPK pathways, including ERK1/2. Given the key role of these pathways in human diseases such as myeloproliferative neoplasms and the challenges associated with modulating such a central pathway, identification of a specific regulator of Mpl signaling poses intriguing questions for future studies on clinical applicability.

Strengths:

Overall, the experiments combine in vitro, in vivo, and ex vivo approaches and appear to have been carefully designed and carried out, with multiple technical and biological replicates where relevant. The authors make a compelling argument for using the Gp1ba-Cre as opposed to the Pf4-Cre system and demonstrate both the dose- and stage-dependent effects of Shp1 and Shp2 on megakaryopoiesis and thrombopoiesis. They find that Shp1 and Shp2 are required in late-stage megakaryocyte maturation and that even low levels of expression compared to baseline are likely sufficient to yield generally normal megakaryocytes. Their findings also lead to specific future directions, such as the mechanism by which Shp1 regulates megakaryopoiesis and thrombopoiesis that is distinct from TPO-mediated signaling.

Weaknesses:

While the experiments have been thoughtfully designed and carried out, there is limited background explanation on relatively complex or niche pathways/mechanisms, such as the relationship between P-selectin, CRP, and PAR4p; the interactions between SFK, Syk, GPVI, and CLEC-2; and TPO, MPL, ERK1/2, AKT, and STAT3, which, while likely intuitive to experts in their respective fields, may be less obvious to a reader approaching this manuscript with a global interest in megakaryopoiesis/thrombopoiesis and thus detract from the impact of the findings.

With regard to the science itself, some of the conclusions feel premature based on the available data.

(1) The section "Aberrant ITAM signaling in Shp1- and Shp2-deficient platelets" is challenging to follow for those not well-versed in ITAM signaling and associated pathways, and may take additional outside reading to follow the conclusion that Syk-dependent signaling is modulated downstream of GPVI and CLEC-2 based on lack of change in Src p-Tyr418, especially considering that Src p-Tyr418 was previously introduced as a measure of SFK rather than Syk. In the introduction, Shp1 is specifically mentioned as a negative regulator of the ITAM/Syk/phospholipase pathway. However, in Figure 4Ai and Bi, Syk phosphorylation/activation in Shp1 knockout cells did not appear to be different from Shp2 knockout cells, and is lower than the control, which is surprising for a negative regulator. It is also not clear why, in the section (Figure 4A-B), there is reduced Syk activation in Shp1 and Shp2 single knockout cells upon CLEC2 stimulation (but apparently not with CRP) when there was no difference in response to CLEC2 (but a difference in response to CRP) in the previous section (Figure 3A, C).

(2) In the section "Reduced Tpo signaling in Shp1/2-deficient MKs," only Western blot data for (p)ERK1/2, AKT, and STAT3 are presented before concluding that decreased ERK1/2 activity is a mechanistic explanation for thrombocytopenia seen in the Shp1/2 double-knockout condition. Such a statement would benefit from additional experiments, such as protein or transcriptional levels of ERK1/2 targets specifically relevant to megakaryopoiesis, such as ETS, FOS, and JUN, to assess the consequences of decreased phosphorylated ERK1/2.

(3) Suggesting that "inhibiting Shp2 will not hav[e] any bleeding consequence in patients" and that Shp2 may be a therapeutic target in myeloproliferative neoplasms when none of these studies have been carried out in a human model is a bold conclusion. There are no data presented on, for example, whether Shp2 inhibition can help reverse the MPL/JAK/STAT pathway in the setting of gain-of-function mutations specifically associated with myeloproliferative neoplasms.

Author response:

The following is the authors’ response to the original reviews.

eLife Assessment

This manuscript provides an important contribution to the field of platelet biogenesis, and the convincing evidence will advance our understanding of signal transduction driving the development of late megakaryopoiesis and platelet reactivity that results in bleeding diathesis. The paper is noteworthy for analyzing two related, either singly or in combination, tyrosine phosphatases in this conditional, stage development gene knockout. Because SHP1 is a negative regulator and SHP2 is an activator, the synergistic effects found in the double knockout were surprising.

We thank the reviewer for acknowledging the importance and novelty of our findings.

Public Reviews:

Reviewer #1 (Public review):

Barré et al. investigated the role of Shp1 and Shp2 in megakaryocytes (MKs) and platelets by conditional knock-out of Shp1, Shp2, or both under the control of the Gp1ba promoter. Deletion of Shp1 and Shp2 in MKs and platelets was almost complete. The Shp1/Shp2 double knock-out mice displayed macrothrombocytopenia and increased bleeding, whereas the single knock-outs did not show significant defects. Platelet function was aberrant in DKOs, but not in single knock-outs, and so was ligand-induced signaling, particularly Syk phosphorylation.

Megakaryocyte maturation was impaired in Shp1/Shp2 DKO mice. Ligand-induced signaling was impaired in Shp2 knock-out and DKO. Ex vivo formation of platelets and in vivo maturation of MKs were impaired in DKO mice. Pharmacological inhibitors of Shp1 and Shp2 had largely similar effects as observed in the single knock-outs. The authors conclude that Shp1 and Shp2 have synergistic functions in the MK/platelet lineage, and that Shp2 may be a potential therapeutic target in myeloproliferative neoplasms.

Strengths:

The data clearly show effects of the Shp1/Shp2 double knock-out on MKs and platelets.

Weaknesses:

There appears to be a discrepancy between the results with the Shp2 single knock-out and the Shp2 inhibitor: the Shp2 knock-out does not affect MKs and platelets, except Erk1/2 signaling, whereas the Shp2 inhibitors appear to affect MK function.

This work is interesting and may have potential from a therapeutic point of view.

Pharmacological effects do not always correlate with congenital anomalies arising for genetic defects. The Shp2 allosteric inhibitors used in our study only inhibit catalytically inactive Shp2, whereas targeted deletion of Ptpn11 results in a loss of total Shp2 expression, including catalytic and non-catalytic related functions, with developmental consequences. Further, Gp1ba-Cre+; Shp2fl/fl megakaryocytes express approximately 22% of normal Shp2 level, which likely also contributes to differences observed between pharmacological inhibition and genetic ablation of Shp2.

We thank the reviewer for recognizing the therapeutic potential of our findings.

Reviewer #2 (Public review):

Summary:

In this manuscript, Barré et al. investigate the roles of the phosphatases Shp1 and Shp2 in the megakaryocyte and platelet lineage using genetic depletion in mice. By employing Gp1ba-Cre-based models, the study builds on the authors' previous work and addresses some limitations associated with earlier Pf4-Cre approaches. The authors report relatively mild alterations in megakaryocyte and platelet parameters in mice lacking either Shp1 or Shp2 alone, whereas combined deletion of both phosphatases results in macrothrombocytopenia, mild bleeding, and impaired GPVI-dependent platelet aggregation accompanied by reduced Syk phosphorylation. The functional platelet defects are linked to reduced expression of GPVI and integrin α2, while thrombocytopenia is associated with impaired megakaryocyte maturation, reduced ploidy, defective proplatelet formation, and altered TPO-dependent Ras/MAPK signaling. Similar effects on megakaryopoiesis are also observed in vitro following treatment with newly developed Shp2 inhibitors.

Strengths and Weaknesses:

The study addresses an important biological question and presents a substantial dataset that could contribute to a better understanding of Shp1 and Shp2 function in platelet biology. However, several aspects of data presentation and interpretation would benefit from additional clarification. In particular, while the authors conclude that single genetic deletion or pharmacological inhibition of Shp1 has a limited impact and that the major phenotypes are specific to combined Shp1/2 deletion or Shp2 inhibition, some of the data suggest more nuanced effects that may warrant further discussion.

We thank the reviewer for raising this point. The manuscript is being revised accordingly, including highlighting the potential role of Shp1 in megakaryopoiesis and thrombopoiesis under steady-state and stressed conditions, requiring more detailed investigation.

Reviewer #3 (Public review):

Summary:

In this manuscript, Barré et al utilize the Gp1ba-Cre transgenic mouse model to build upon previous findings in a Pf4-Cre system to investigate the effects of individual and combined Shp1 and Shp2 deletion in megakaryocytes and platelets. They report decreased megakaryocyte maturation, macrothrombocytopenia, and increased bleeding primarily in association with the Shp1/Shp2 double-knockout condition. The authors further show that this phenotype appears to be driven primarily by Shp2 and implicate dysregulation of Mpl signaling and downstream Ras/MAPK pathways, including ERK1/2. Given the key role of these pathways in human diseases such as myeloproliferative neoplasms and the challenges associated with modulating such a central pathway, identification of a specific regulator of Mpl signaling poses intriguing questions for future studies on clinical applicability.

We thank the reviewer for acknowledging the importance and novelty of our findings.

Strengths:

Overall, the experiments combine in vitro, in vivo, and ex vivo approaches and appear to have been carefully designed and carried out, with multiple technical and biological replicates where relevant. The authors make a compelling argument for using the Gp1baCre as opposed to the Pf4-Cre system and demonstrate both the dose- and stagedependent effects of Shp1 and Shp2 on megakaryopoiesis and thrombopoiesis. They find that Shp1 and Shp2 are required in late-stage megakaryocyte maturation and that even low levels of expression compared to baseline are likely sufficient to yield generally normal megakaryocytes. Their findings also lead to specific future directions, such as the mechanism by which Shp1 regulates megakaryopoiesis and thrombopoiesis that is distinct from TPO-mediated signaling.

Weaknesses:

While the experiments have been thoughtfully designed and carried out, there is limited background explanation on relatively complex or niche pathways/mechanisms, such as the relationship between P-selectin, CRP, and PAR4p; the interactions between SFK, Syk, GPVI, and CLEC-2; and TPO, MPL, ERK1/2, AKT, and STAT3, which, while likely intuitive to experts in their respective fields, may be less obvious to a reader approaching this manuscript with a global interest in megakaryopoiesis/thrombopoiesis and thus detract from the impact of the findings.

We thank the reviewer for raising this point. The manuscript is being revised to better explain the rationale and molecular mechanisms linking these pathways and functions.

With regard to the science itself, some of the conclusions feel premature based on the available data.

(1) The section "Aberrant ITAM signaling in Shp1- and Shp2-deficient platelets" is challenging to follow for those not well-versed in ITAM signaling and associated pathways, and may take additional outside reading to follow the conclusion that Syk-dependent signaling is modulated downstream of GPVI and CLEC-2 based on lack of change in Src p-Tyr418, especially considering that Src p-Tyr418 was previously introduced as a measure of SFK rather than Syk. In the introduction, Shp1 is specifically mentioned as a negative regulator of the ITAM/Syk/phospholipase pathway. However, in Figure 4Ai and Bi, Syk phosphorylation/activation in Shp1 knockout cells did not appear to be different from Shp2 knockout cells, and is lower than the control, which is surprising for a negative regulator. It is also not clear why, in the section (Figure 4A-B), there is reduced Syk activation in Shp1 and Shp2 single knockout cells upon CLEC2 stimulation (but apparently not with CRP) when there was no difference in response to CLEC2 (but a difference in response to CRP) in the previous section (Figure 3A, C).

We thank the reviewer for raising these important points. The manuscript is being revised accordingly, including clarifying the roles of SFKs, Shp1 and Shp2 in the ITAM-Syk-PLCγ2 signaling pathway.

Briefly, SFKs are essential for phosphorylating ITAMs, allowing SH2-dependent docking of Syk. Reduced reactivity of Shp1/2 DKO platelets to CRP and collagen is likely due to downregulation of the ITAM-containing GPVI-FcR γ-chain complex and integrin α2 subunit, and concomitant reduction in Syk phosphorylation.

However, the marginal albeit significant reduction in Syk phosphorylation downstream of CLEC-2 in Shp1 and Shp2 KO platelets was not determined and was insufficient to impact CLEC-2-mediated platelet aggregation under the conditions tested.

Differences in the stoichiometry and docking of Syk to phosphorylated GPVI-FcR γ-chain and CLEC-2 likely contribute to the differences in platelet reactivity and Syk phosphorylation downstream of the two receptors in the absence of Shp1 and Shp2.

(2) In the section "Reduced Tpo signaling in Shp1/2-deficient MKs," only Western blot data for (p)ERK1/2, AKT, and STAT3 are presented before concluding that decreased ERK1/2 activity is a mechanistic explanation for thrombocytopenia seen in the Shp1/2 doubleknockout condition. Such a statement would benefit from additional experiments, such as protein or transcriptional levels of ERK1/2 targets specifically relevant to megakaryopoiesis, such as ETS, FOS, and JUN, to assess the consequences of decreased phosphorylated ERK1/2.

We thank the reviewers for these constructive comments. Further experiments are being planned to determine the biological and transcriptional consequences of reduced ERK1/2 phosphorylation during megakaryopoiesis and thrombopoiesis.

(3) Suggesting that "inhibiting Shp2 will not have any bleeding consequence in patients" and that Shp2 may be a therapeutic target in myeloproliferative neoplasms when none of these studies have been carried out in a human model is a bold conclusion. There are no data presented on, for example, whether Shp2 inhibition can help reverse the MPL/JAK/STAT pathway in the setting of gain-of-function mutations specifically associated with myeloproliferative neoplasms.

This conclusion is being tempered in the revised manuscript. Genetic- and pharmacological-based approaches will be used to establish the therapeutic potential of inhibiting Shp1 and Shp2 in mouse models of MPN, including Jak2 gain-of-function mice. Bleeding and thrombotic complications of inhibiting Shp1 and Shp2 will be explored as part of these studies.

Recommendations for the authors:

Reviewing Editor Comments:

Altogether, we feel that this is an important study for those in the fields of hematology or signal transduction. Your important study characterizes the roles in late megakaryopoiesis and platelet biogenesis of single or combined conditional deletion of two tyrosine phosphatases, Shp1 and Shp2. Strengths include technical advances in single and combined deletions, the somewhat surprising results of synergy between the two phosphatases, focusing on the critical stage of late megakaryopoiesis, and clinical implications in bleeding diathesis.

Weaknesses are mostly minor, but the numerous points raised by reviewer 3 need to be addressed and typographical errors corrected. Further discussion should include the relevance or dissimilarity in megakaryopoiesis and platelet biogenesis between murine and human blood health and disease. Since SHP1 is a negative regulator and SHP2 is a positive activator, additional discussion about how they coordinate and fine-tune ("nuanced") signal transduction in TPO- or GPVI-induced signaling in an explicitly stated pathway.

We invite you to respond to the critiques and submit a revised manuscript.

Sincerely,

Seth Corey, MD MPH

We thank the editor for the positive evaluation of our study and for highlighting its relevance to the fields of haematology and signal transduction. We have carefully addressed all comments raised by Reviewer 3 and corrected typographical errors throughout the manuscript.

As suggested, we expanded the Discussion to better address the relevance of our murine findings to human megakaryopoiesis and platelet biogenesis. While our study relies on mouse models, key components of TPO/MPL signaling and platelet production are conserved between mice and humans, although differences in megakaryocyte maturation dynamics and platelet biology are acknowledged and now discussed.

We also clarified the coordinated roles of Shp1 and Shp2 in signaling. Although Shp1 generally acts as a negative regulator and Shp2 as a positive mediator of signal transduction, our results suggest that they function in a complementary manner to optimize signaling downstream of TPO/MPL and GPVI pathways, thereby ensuring appropriate regulation of late megakaryopoiesis, platelet production and activation.

These additional considerations have been incorporated into the revised manuscript to provide a clearer conceptual framework for how Shp1 and Shp2 cooperate to regulate platelet biogenesis.

Reviewer #1 (Recommendations for the authors):

(1) The effects of the Shp1/Shp2 DKO are clear, but the effect of the Shp2 single knock-out is less clear on all parameters that were tested. The exception is ERK1/2 phosphorylation, which was reduced in the Shp2 knock-out as well as the Shp1/Shp2 DKO. Why do the authors conclude that Shp2 may be a potential therapeutic target, while the data show that knock-out of Shp1 and Shp2 is required for the observed effects?

We agree that the most pronounced phenotypes were observed in the Shp1/Shp2 DKO. However, Shp2 single knock-out consistently reduced ERK1/2 phosphorylation, indicating that Shp2 contributes to MPL downstream signaling in megakaryocytes. The absence of a strong phenotype in Shp2 single knock-out may be due to residual Shp2 protein. However, given the established role of the Shp2–ERK pathway in megakaryopoiesis and the observation that pharmacological Shp2 inhibition significantly affected MK ploidy, proplatelet formation, and ERK1/2 phosphorylation, our data support a contribution of Shp2 to these processes and suggest it as a potential therapeutic target.

(2) Inhibitors of Shp1 and Shp2 had largely similar effects as Shp1 and Shp2 single knock-outs, respectively. The effect of Shp2 knock-out on MK ploidy is not clear, cf. Figure 5Ai (no effect) and Figure 5Aii (reduction, which is not significant), whereas a clear and significant effect was reported for the Shp1/Shp2 DKO. In contrast, in Figure 7Ciii, the Shp2 inhibitors SHP099 and RMC-4550 clearly affect MK ploidy and the percentage of MKs forming proplatelets. The discrepancy between the effect of Shp2 knock-out and Shp2 inhibitors suggests that the inhibitors may affect other targets. The authors should consider using the Shp2 inhibitors on the Shp2 knock-out to prove or disprove that the effects of the Shp2 inhibitors are mediated exclusively by Shp2.

Pharmacological inhibition does not necessarily phenocopy genetic deletion. The allosteric Shp2 inhibitors used in our study (SHP099 and RMC-4550) stabilize Shp2 in an inactive conformation and inhibit its catalytic activity, whereas Ptpn11 deletion results in complete loss of the Shp2 protein, including both catalytic and scaffolding functions. These mechanistic differences may lead to distinct biological outcomes and could explain the discrepancy observed between Shp2 knockout and inhibitor treatments.

(3) Since the most profound effects were found in the Shp1/Shp2 DKO, it would be interesting to use combinations of the Shp1 and Shp2 pharmacological inhibitors to mimic the effect of the Shp1/Shp2 DKO.

We thank the reviewers for these constructive comments. Further experiments are indeed being planned to use combinations of the Shp1 and Shp2 pharmacological inhibitors to mimic the effect of the Shp1/2 DKO.

Reviewer #2 (Recommendations for the authors):

Major points:

(1) Additional details on the strategy used to isolate megakaryocyte progenitors from mouse bone marrow would improve clarity, including sorting approach, gating strategy, and assessment of population purity.

We thank the reviewer for this suggestion. We have now expanded the Methods section to provide a more detailed description of the strategy used to isolate megakaryocyte progenitors from mouse bone marrow.

Briefly, bone marrow cells were first enriched for hematopoietic progenitors and stained with antibodies against lineage markers and megakaryocyte-associated markers. Megakaryocyte progenitors were then isolated by flow cytometric sorting based on established surface marker combinations, including c-Kit and CD41 expression. The gating strategy excluded lineage-positive cells and debris before selecting the progenitor population of interest.

(2) Platelet GPVI expression appears reduced not only in Shp1/2 double-knockout mice but also, to some extent, in single Shp1- or Shp2-deficient models. A more detailed quantitative comparison and discussion would be helpful.

We thank the reviewer for this observation. Although the most pronounced reduction in GPVI surface expression was observed in Shp1/Shp2 double knock-out platelets, minor variations may appear in the single knock-out models. To address this, we performed additional statistical analyses comparing WT platelets with each single knock-out genotype. These analyses did not reveal any significant statistical differences in GPVI expression between WT and either Shp1- or Shp2-deficient platelets, indicating that the apparent variations fall within the range of biological variability.

(3) The aggregation traces shown in Figures 3A and 3B would benefit from clarification regarding their representativeness relative to the corresponding quantitative analyses.

We thank the reviewer for this comment. The aggregation traces in Figures 3A and 3B represent experiments selected from independent replicates included in the quantitative analysis. The figure legends have been revised to clarify that these traces are representative of the experiments summarized in the quantification panels, which include data from multiple independent mice.

(4) In several experiments, statistical significance may be influenced by differences in sample size across genotypes (e.g., Figures 2Ci, 3Ai, 3Di, and 6Ai). Using comparable numbers of replicates would strengthen the interpretation.

We appreciate the reviewer’s attention to statistical rigour. The differences in sample size between genotypes reflect the availability of animals from the different breeding cohorts. Importantly, all statistical analyses were performed using appropriate tests that account for unequal sample sizes. The observed differences remain consistent across independent experiments.

(5) The rationale for assessing only P-selectin exposure following CRP and PAR4p stimulation is not fully explained. Including integrin αIIbβ3 activation, or clarifying its exclusion, would provide a more complete assessment of platelet activation.

We thank the reviewer for this suggestion. P-selectin exposure was used as a primary readout because it provides a robust measure of α-granule secretion downstream of GPVI and PAR signaling. Integrin αIIbβ3 activation was not assessed in these experiments because platelet aggregation assays were performed in parallel, which already provide a functional readout of integrin activation, as aggregation requires αIIbβ3 engagement. Nonetheless, we agree with the reviewer that direct measurement of integrin activation (e.g., fibrinogen binding) would provide complementary information and will be considered in future studies.

(6) Figure 3Dii is described as an aggregation assay, although it appears to report P-selectin exposure; this distinction should be clarified.

We thank the reviewer for identifying this inconsistency. Figure 3Dii reports indeed P-selectin exposure measured by flow cytometry, rather than platelet aggregation. We have corrected the description in the Results section.

(7) The suggestion of compensatory extramedullary hematopoiesis based on splenomegaly would be strengthened by immunophenotypic analysis of splenic hematopoietic progenitor populations.

We appreciate this important suggestion. In the current study, the evidence for possible compensatory extramedullary hematopoiesis is mainly based on the splenomegaly observed in Shp1/2 DKO mice. We agree that detailed immunophenotypic analysis of splenic hematopoietic progenitors would provide additional mechanistic insight; however, this was beyond the scope of the present study, which focuses on the intrinsic role of Shp1 and Shp2 in the megakaryocyte and platelet lineage. We have therefore revised the Discussion to present this interpretation more cautiously and to indicate that further studies will be required to determine whether splenic hematopoiesis contributes to compensatory platelet production in this model.

(8) In Figure S3, differences in platelet recovery kinetics among genotypes appear evident. Clarification of the statistical tests used to assess these differences would be useful.

We thank the reviewer for this comment. Platelet recovery kinetics were analyzed using two-way ANOVA with appropriate post hoc tests. No statistically significant differences between genotypes were observed. These details have been added to the Methods and figure legend for clarity.

Reviewer #3 (Recommendations for the authors):

Overall, the manuscript suffers from multiple typographical and grammatical errors that distract from the data being presented.

We have carefully revised the manuscript to correct typographical and grammatical errors throughout, improving clarity and readability.

(1) Figure S1: I believe this should be referenced in the first paragraph of the results section.

We have now referenced the Supplemental Figure S1 in the first paragraph of the results section as suggested.

(2) Figure 2A: Although the individual points for the replicates are informative, they do make it difficult to appreciate the SEM, and to my eye it appears that, for example, there may not be a difference between Shp2 and Shp1/2 or that there may be a difference between Shp1 and Shp1/2 in (ii), as Table S2 suggests. In other words, it seems that the increased MPV (as well as the leukocyte phenotype) may be driven by the knockout of Shp2; are there statistical analyses that could be performed to show that the increased MPV is specific to the double knockout?

We thank the reviewer for this comment. Despite the slightly higher MPV observed in Shp2 single knockouts, statistical analysis using one-way ANOVA, which is appropriate for comparing means across multiple independent groups, and taking all individual data points into account, revealed no significant differences between Shp2 or Shp1 single KO and the Shp1/2 DKO.

(3) Figure 2Bi: Is this missing a statistical significance bar, or was there no significant difference in cumulative bleeding time between the conditions? If the latter, this should be clarified in the main text (although the specific sentence regarding bleeding time only claims "mildly prolonged," the preceding sentence indicates "significant increase in bleeding").

Thank you for this comment. There was no statistically significant difference in cumulative bleeding time between the groups. We have now modified the text accordingly to clarify this point and to indicate that, while bleeding time was not significantly different, blood loss was significantly increased in Shp1/2 DKO mice.

(4) Figure 2Ci: What was the extent (statistically) of GPVI reduction in the Shp1 and Shp2 single knockout mice compared to the control? It seems that although there was no change in alpha2 expression in the single-knockout conditions, the contributions of Shp1 and Shp2 loss may be additive on GPVI (although I acknowledge that this is not necessarily borne out in Figure 3Ai).

Thank you for this comment. After reanalyzing the data using an appropriate statistical test (one-way ANOVA followed by Tukey’s post hoc test), we found that GPVI expression is significantly reduced in both Shp1 and Shp2 single knockout platelets compared with controls. However, this reduction did not result in detectable functional consequences on platelet aggregation, as shown in Figure 3Ai.

(5) Figure 3Ai: It seems that the individual replicates for the Shp1/2 double knockout cluster in two populations, extreme non-responders and arguably normal responders to CRP. Are there any biological or technical explanations for this?

We thank the reviewer for this observation. We agree that the distribution of individual replicates in the Shp1/2 DKO group suggests the presence of two subpopulations, with some samples showing markedly impaired aggregation and others retaining near-normal responsiveness to CRP. While all experiments were performed under standardized conditions, subtle differences in platelet preparation, agonist sensitivity, or assay timing could also contribute to dispersion within this group. Importantly, despite this variability, the overall trend indicates a significant reduction in aggregation in the Shp1/2 DKO condition compared to controls, supporting a critical and partially redundant role for Shp1 and Shp2 in GPVI-mediated platelet activation.

(6) "Aberrant functional responses of Shp1/2-deficient platelets": It may be helpful, in the last paragraph of this section, to briefly explain the relationship between P-selectin, CRP, and PAR4p. If short on space/words, the introduction likely does not need an explanation of platelet function and definitions of megakaryopoiesis and thrombopoiesis.

We thank the reviewer for this suggestion. We have revised the last paragraph to clarify that P-selectin surface expression reflects α-granule secretion following platelet activation. We now specify that CRP activates platelets via GPVI signaling, whereas PAR-4 peptide signals through thrombin receptors, providing context for the differential responses observed in Shp1/2-deficient platelets.

(7) "Aberrant ITAM signaling in Shp1- and Shp2-deficient platelets": Is there a cartoon figure panel that could be added to clarify how SFK (which, as an aside, is not defined as an acronym), Syk, GPVI, CLEC-2 receptor, Shp1, and Shp2 are interrelated? In addition to the comments left in the public review, I was perplexed by Figure 4Bi, as the band for the Shp1/2 double knockout condition appears to be stronger than the other 3 conditions, but this is not what is depicted in the bar graph on the right.

We thank the reviewer for this helpful comment. We have now added a schematic cartoon (new Figure 8) to clarify the relationships between SFKs, Syk, GPVI, and the regulatory roles of Shp1 and Shp2. All acronyms, including SFK, are now defined at first mention to improve accessibility.

Regarding Figure 4Bi, we appreciate this observation. The apparent discrepancy between the representative blot and the quantification reflects variability across experiments. The bar graph represents the average of independent replicates.

(8) I would also recommend considering reshuffling the panels in Figure 4 so that the 2 assays measuring Syk phosphorylation and the 2 assays measuring Src phosphorylation are next to each other, as opposed to grouped by agonist. They should also be presented in the order of the text, which states that SFK activation was measured via Src before mentioning Syk (but the data are presented in reverse).

We thank the reviewer for this suggestion. We have reorganized Figure 4 so that the panels measuring Src and Syk phosphorylation are presented together and, in the order, described in the text. The manuscript text has also been updated accordingly to match the revised figure layout.

(9) GPVI overexpression experiments in these megakaryocytes or, conversely, Syk inhibition in control cells, to reverse or recapitulate the phenotype, respectively, may be additionally informative.

We thank the reviewer for this suggestion. We agree that modulating GPVI or Syk activity could provide additional mechanistic insight. While these experiments were beyond the scope of the current study, we plan to explore GPVI overexpression and Syk inhibition in follow-up studies to further validate the pathway’s role in the observed phenotype.

(10) "Reduced Tpo signaling in Shp1/2-deficient MKs": In addition to the comments left in the public review, I would suggest moving this section to after "Defective proplatelet formation and MK maturation in Shp1/2-deficient mice" so that the 2 sets of proplatelet and ploidy data are consecutively presented.

We thank the reviewer for this helpful suggestion. We have now revised the manuscript accordingly by reorganizing both the text and figures. The ploidy and proplatelet formation data are now presented together in Figure 5, followed by the Tpo signaling data in Figure 6, improving the overall flow and clarity of the results section.

(11) Figure 6Cii: Why does Shp1 add up to >100%?

The reason the Shp1 bar exceeds 100% is due to how the data were quantified and normalized. Each segment represents the mean from separate experiments. Stacking these means can exceed 100% because the sum of averages is not equal to the average of the total.

(12) Figure 7D: How do you reconcile these findings of impaired AKT phosphorylation with the addition of a Shp2 inhibitor but no change with Shp2 knockout (Figure 5C)? Would you attribute it to the residual Shp1 and Shp2 in the Cre-Lox MKs?

Pharmacological effects do not always correlate with congenital anomalies arising for genetic defects. The Shp2 allosteric inhibitors used in our study only inhibit catalytically inactive Shp2, whereas targeted deletion of Ptpn11 results in a loss of total Shp2 expression, including catalytic and non-catalytic related functions, with developmental consequences. Further, Gp1ba-Cre+; Shp2fl/fl megakaryocytes express approximately 22% of normal Shp2 level, which likely also contributes to differences observed between pharmacological inhibition and genetic ablation of Shp2.

Version 2

Abstract

The Src homology 2 (SH2) domain-containing non-transmembrane protein-tyrosine phosphatases 1 and 2 (Shp1 and Shp2) have been implicated in regulating signaling from a variety of receptors and cell types, including the thrombopoietin (Tpo) receptor Mpl in megakaryocytes (MKs) and platelets. We previously showed that deletion of Shp1 and Shp2 in the MK/platelet lineage in mice using the Pf4-Cre transgene/loxP system impairs megakaryopoiesis and thrombopoiesis. However, we also observed unexpected phenotypes including a motheaten-like phenotype in Shp1-deficient mice and severe myelofibrosis in mice lacking both phosphatases. To determine whether these were lineage-specific effects, we utilized the Gp1ba-Cre transgenic mouse to delete loxP-flanked Shp1 and Shp2 in mice. Bone marrow-derived MKs from these mice expressed approximately 20-25% of Shp1 and Shp2, whereas platelets contain 5-10% of each phosphatase compared with controls. Minor MK/platelet defects were observed in mice lacking either Shp1 or Shp2 alone, however mice lacking both Shp1 and Shp2 exhibited macrothrombocytopenia, mild bleeding following tail injury, and impaired GPVI-mediated platelet aggregation and Syk phosphorylation, associated with reduction GPVI and integrin α2 subunit expression. Reduced Shp1 and Shp2 expression resulting in a significant reduction in ploidy, a block in MK maturation and proplatelet-producing MKs. Tpo-mediated Ras/MAPK signaling was reduced in Shp1/2-deficient MKs. Treatment of MKs with structurally distinct Shp2 allosteric inhibitors recapitulated key aspects of the Shp2-deficient phenotype, including aberrant megakaryopoiesis and reduced Mpl signaling. Our study highlights the synergistic functions of Shp1 and Shp2 in the MK/platelet lineage, and identifies Shp2 as a potential therapeutic target in myeloproliferative neoplasms.

Introduction

Platelets, or thrombocytes are small, anucleate blood cells derived from megakaryocytes (MKs) that are essential for hemostasis and thrombosis.1 MK development, or megakaryopoiesis and platelet production, or thrombopoiesis are tightly regulated processes that take place in the bone marrow (BM), spleen and lungs, but the molecular mechanisms regulating these inter-related processes remain incompletely defined.2,3 The primary driver of megakaryopoiesis, which encompasses differentiation of hematopoietic stem and progenitor cells (HSPCs) to MKs, proliferation, endomitosis and maturation is the cytokine thrombopoietin (Tpo) produced by the kidneys, liver and BM acting on its receptor, myeloproliferative leukemia protein (Mpl) expressed throughout the MK/platelet lineage.4 Tpo does not however explain which MKs undergo proplatelet formation and thrombopoiesis, which occurs when MKs come in contact with sinusoidal blood vessels. Several other cytokines and growth factors, extracellular matrix (ECM) proteins and chemokines are also required for optimal megakaryopoiesis and thrombopoiesis, as evidenced in mice lacking either Tpo or Mpl, which still produce MKs and platelets, albeit at much lower levels than normal mice.5,6

Tpi signaling through its receptor Mpl is a central regulator of megakaryopoiesis and platelet production. Upon ligand binding, Mpl undergoes conformational changes that activate three primary signaling pathways : Janus kinase 2 (JAK2)/Signal transducer and activator of transcription (STAT), Ras/mitogen-activated protein kinase (MAPK) and phosphatidylinositol 3-kinase (PI3K)/AKT, all of which culminate in nuclear translocation of signaling/transcription factors and gene regulation.7 These pathways act in a coordinated manner to control both the expansion and functional maturation of megakaryocytes, thereby linking extracellular Tpo stimulation to the cellular processes governing thrombopoiesis. Cytoskeletal remodeling via the RhoA/Rho-associated protein kinases (ROCK) pathway is also an integral part of megakaryopoiesis and thrombopoiesis, notably during endomitosis8, but the molecular link with the Mpl receptor is undefined. Clonal gain-of-function mutations in the Mpl/JAK2/STAT pathway in HSPCs are the most common causes of myeloproliferative neoplasms (MPN)9, characterized by increase MK/platelet production (essential thrombocythemia, ET), increased red blood cell production (polycythemia vera, PV) and associated fibrotic deposits (primary myelofibrosis, PMF).10–12 Notably, the valine 617 phenylalanine (V617F) mutation in the pseudo-kinase domain of JAK2, resulting in increased catalytic active is the most prevalent, found in 60% of MPN cases, followed by C-terminal deletions in calreticulin, resulting in the release of truncated calreticulin and it acting as a ligand of Mpl in 30% of the cases, and mutations in Mpl resulting in increased expression in <5% of the cases.13 JAK2 and MAPK kinase (MEK) inhibitors are effective in some instances of MPN14, however there is a tendency of resistance and relapse, requiring alternative and improved therapeutics.

The Src homology 2 (SH2) domain-containing non-transmembrane protein-tyrosine phosphatases (PTPs) 1 and 2 (Shp1 and Shp2) are involved in hematopoietic cell differentiation, proliferation, function and survival, including megakaryopoiesis and thrombopoiesis.15 Intra-and inter-molecular interactions of the tandem SH2 domains with the catalytic domain and phospho-tyrosine binding partners regulate the activity and compartmentalization of both phosphatases, respectively. Tandem phospho-tyrosine residues in the C-terminal tails of Shp1 and Shp2 are implicated in substrate binding and regulation.16 Despite high structural similarities, the two phosphatases have distinct biological functions. Shp1, encoded by the PTPN6 gene, is primarily expressed in hematopoietic cells and is a negative regulator of immune receptor signaling, particularly the immunoreceptor tyrosine-based activation motif (ITAM)/Spleen tyrosine kinase (Syk)/phospholipase pathway.17 Shp2, encoded by the PTPN11 gene, is ubiquitously expressed and is a positive regulator of cytokine and growth factor signaling, specifically the Ras/MAPK and PI3K/AKT pathways.18 However, their functions go beyond these pathways and are context dependent. Downstream substrates and effectors of both phosphatases remain elusive.

We previously showed that targeted deletion of Shp1 and Shp2 in the MK/platelet lineage using the Pf4-Cre transgene/loxP system impairs megakaryopoiesis and thrombopoiesis in mice.15 However, we also found that these mice exhibited unexpected, non-MK/platelet-related defects, including a motheaten-like phenotype in Shp1-deficient mice, and severe myelofibrosis in mice lacking both phosphatases. Intriguingly, the motheaten-like phenotype was lost in Shp1/2 double-deficient mice, suggesting that Shp1 and Shp2 have opposing effect in this pathology. To determine whether these were in fact MK/platelet lineage-specific effects, we utilized the Gp1ba-Cre transgenic mouse to delete loxP-flanked Shp1 and Shp2 in mice. In contrast to Pf4-Cre transgenic mice, conditional deletion of Shp1 or Shp2 using the Gp1ba-Cre transgene had little if any detectable effect on megakaryopoiesis or thrombopoiesis in these mice under steady state conditions. However, deletion of both Shp1 and Shp2 using Gp1ba-Cre resulted in macrothrombocytopenia and aberrant MK development and maturation, due to a block in Mpl-mediated Ras/MAPK signaling. GPVI-induced platelet aggregation was also impaired due to reduced GPVI and α2 integrin subunit expression. Similar effects were seen with the structurally-distinct allosteric Shp2 inhibitors, SHP099 and RMC-4550.19 Findings demonstrate synergistic effects of deleting Shp1 and Shp2 in the MK/platelet lineage in mice by disrupting distinct signaling pathways, and highlight Shp2 as a potential therapeutic target in MPN.20,21

Materials and Methods

Mouse models

All mice used were on a C57BL/6 background. Ptpn6fl/fl, Ptpn11fl/fl and Gp1ba-Cre+/KI mice were generated, as previously described.22–24 MK/Platelet specific Shp1 and Shp2 knockout (KO) mice were generated by crossing Ptpn6fl/fl and Ptpn11fl/fl mice with the Gp1ba-Cre transgenic deleter mouse Gp1ba-CreKI/+ mice. Gp1ba-Cre KI/+ were used as control mice (CT) (Supplemental Figure S1). All animal experiments were conducted in accordance with the CREMEAS Committee on the Ethics of Animal Experiments of the University of Strasbourg (Permit Number: E67-482-10, Project approval number: APAFIS#28221-2020111714459066).

Inhibitors

SHP099 and RMC-4550 Shp2 allosteric inhibitors were purchased from MedChemExpress. M029 is a covalent allosteric Shp1 inhibitor (Z.Q. and Z.Y.Z, manuscript in preparation) and F2Ac is a reversible active site directed Shp1 inhibitor (J.M. and Z.Y.Z., manuscript in preparation.

Platelet preparation and functional assays

Blood was collected from the aorta of anesthetized mice into 1/10 acid-citrate-dextrose (ACD) anticoagulant. Platelet counts were measured from peripheral blood samples using an automated hematology analyzer (Element HT5 counter, Heska company). Washed platelets were prepared as previously described.25 Platelet counts were normalized and used for aggregation (2 x 108/ml) or biochemical analysis (5 x 108/ml). Platelet aggregation was measured using the lumi-aggregometer APACT®4004. Surface glycoprotein expression was measured in whole blood by flow cytometry using fluorescein isothiocyanate (FITC)- conjugated antibodies. Resting and activated platelets were fixed and stained with anti-P-selectin antibody.

Megakaryocyte culture and functional assays

Mature MKs from mouse BM were cultured and analysed, as previously described.26 BM cells were isolated from femora, tibiae and iliac crests of C57BL6 mice for pharmacological study and Gp1ba-Cre mice for other experiments by flushing the bones with Dulbecco’s Modified Eagle Medium (DMEM) and successively passed through 21-, 23-, and 25-gauge needles to obtain a single-cell suspension. After centrifugation, BM cells were subjected to immunomagnetic lineage-negative (Lin−) selection (StemCell Technologies) and were cultured for 3 days at 37°C 5% CO2 at a concentration of 1x106 cells/well in DMEM supplemented with 10% Fetal Bovine Serum (FBS), 50 ng/ml Tpo, 100 U/ml Hirudin, 1% PSG.26 MK ploidy was assessed after 3 days of culture, using FITC-coupled anti-CD41 antibody and DNA staining with propidium iodide (PI). Samples were acquired using the BD LSRFortessa X-20 flow cytometer and analyzed using BD FACSDiva software. The ability of MKs to form proplatelets was assessed in vitro, and ex vivo using the explant assay, as previously described.27

Immunoblotting

Platelet and MK whole cell lysates (WCLs) were prepared and analyzed by automated capillary-based immunoassay (ProteinSimple Jess), prepared according to manufacturer’s instructions, as previously described.28 The High Dynamic Range profile was used for chemiluminescent and fluorescent multiplexing signal detection. Optimized antibody dilutions and sample concentrations used are provided in Supplemental Table 1.

Immunohistochemistry

Spleens from KO and litter-matched CT mice were fixed in buffered formalin and embedded in paraffin. Sections (5 μm) were H&E stained and examined by light microscopy using a 40× objective.

Electron microscopy

Maturation stage of MKs was assessed by electron microscopy (EM). BM samples obtained by flushing mouse femora with 0.1 M sodium cacodylate buffer were fixed in 2.5% glutaraldehyde and embedded in Epon as previously described.29 Thin sections were stained with uranyl acetate and lead citrate, and examined under a JEOL 2100Plus transmission electron microscope at 120 kV (Jeol, Tokyo, Japan). MKs at stages I, II and III were identified using distinct ultrastructural characteristics: stage I, a cell 10–15 μm in diameter with a large nucleus; stage II, a cell 15–30 μm in diameter containing platelet-specific granules; stage III, 40 µm, a MK containing a well-developed demarcation membrane system defining cytoplasmic territories and a peripheral zone. Samples from at least three mice of each genotype were examined in each case.

Platelet recovery assay

An intraperitoneal injection of anti-GPIbα antibody (Emfret) at 2 mg/kg is performed at D0. Blood samples were collected and platelet counts measured prior to injection, and at D5, D10, D15 and D20.30 Platelet recovery kinetics were analyzed using two-way ANOVA with appropriate post hoc tests.

HSPC colony-forming unit (CFU) assay

Cells are obtained by flushing and were resuspended in IMDM and plated at a density of 1×10⁴ cells per dish in methylcellulose-based medium (MethoCult™ GF M3434, Stemcell Technologies) containing appropriate cytokines to support multilineage hematopoietic colony formation. Cultures were incubated at 37 °C with 5% CO₂ in a humidified incubator for 10–14 days. Colonies were scored under an inverted microscope and classified based on morphological criteria into CFU-G (granulocyte), CFU-M (macrophage), CFU-GM (granulocyte-macrophage), BFU-E (erythroid), and CFU-GEMM (granulocyte, erythrocyte, macrophage, megakaryocyte). Only colonies containing ≥50 cells were counted. Results were expressed as the mean colony number ± SEM from n = 3 independent experiments.

Statistical analysis

Statistical analysis was performed using GraphPad Prism version 9 software. Data were expressed as mean ± standard error mean (SEM). Statistical significance was analyzed by one-or two-way ANOVA followed by the appropriate post hoc test, or Kruskal-Wallis test for nonparametric data, a Bonferroni correction was applied for multiple comparisons, as indicated in figure legends. P-values < 0.05 were considered statistically significant.

Results

Deletion of Shp1 and Shp2 in the MK/platelet lineage

To validate the efficiency of protein ablation of Shp1 and Shp2 proteins in MKs/platelets using the Gp1ba-Cre transgenic mouse strain (Supplemental Figure S1), we employed a quantitative capillary-based immunoassay. This was performed on cells derived from Ptpn6fl/fl;Gp1ba-Cre+/KI (Shp1 KO), Ptpn11fl/fl;Gp1ba-Cre+/KI (Shp2 KO) and Ptpn6fl/fl;Ptpn11fl/fl;Gp1ba-Cre+/KI double-knockout (DKO) mice, with Gp1ba-Cre+/KI littermates serving as controls.

We firstly evaluated Shp1 and Shp2 levels in MK-sorted progenitors (MKPs) in order to obtain a pure population of matures MKs. Under these conditions, we quantified protein loss of Shp1 and Shp2 in MKs, as shown in Figure 1Ai-iii, with significant 70% and 82% reduction of Shp1 protein in Shp1 KO and Shp1/2 DKO mice, respectively and 78% and 75% reduction of Shp2 protein in Shp2 KO and Shp1/2 DKO mice, respectively compared to controls.

Protein expression of Shp1 and Shp2 in megakaryocytes and platelets from Gp1ba-Cre mice.

(i) Representative blots of Shp1 and Shp2 expression levels from CT, Shp1, Shp2 and Shp1/2 DKO mice by capillary immunoassays with the respective antibodies in (A) sorted-MK progenitors and (B) washed platelets. Percentage of residual level of (ii) Shp1 and (iii) Shp2 in Shp1 KO (blue), Shp2 KO (green) and Shp1/2 DKO (red) mice. n = 3-6 mice per genotype. Mean ± SEM; one way-ANOVA; ** P < 0.01, *** P < 0.001.

We next evaluated protein levels in washed platelets and found efficient deletion in Shp1 KO and Shp2 KO platelets, with a significant 90% and 92% reduction of Shp1 protein in Shp1 KO and Shp1/2 DKO mice respectively, and 95% and 96% reduction of Shp2 protein in Shp2 KO and Shp1/2 DKO mice respectively, compared to controls (Figure 1Bi-iii). In addition, there was no upregulation of Shp1 in Shp2-deficient MKs/platelets and vice versa. These results indicate that the Gp1ba-Cre transgene efficiently deleted loxP-flanked Ptpn6 and Ptpn11, resulting in the significant loss of Shp1 and Shp2 in MKs/platelets.

Macrothrombocytopenia and increased bleeding in Shp1/2 DKO mice

Deletion of Shp1 or Shp2 using the Gp1ba-Cre-mouse strain did not affect platelet counts and volumes as shown in Figure 2Ai-ii and Supplemental Table 2. However, Shp1/2 DKO mice were macrothrombocytopenic, with a 40% reduction in platelet count and 15% increase in platelet volume (Figure 2Ai-ii). Monocyte, neutrophil and eosinophil counts were marginally increased in Shp1/2 DKO mice, suggesting possible immune responses. (Supplemental Table 2).

Macrothrombocytopenia and mild increased bleeding in Shp1/2 DKO mice.

(A) Platelet counts (i) and platelet volumes (ii) of control (CT) (n=75), Shp1 KO (n=21), Shp2 KO (n=29) and Shp1/2 DKO (n=46) mice. Mean ± SEM; one way-ANOVA; *** P < 0.001. (B) (i) Cumulative bleeding time was recorded (ii) and the volume of blood loss was measured over 30 minutes. Mean ± SEM; one way-ANOVA; * P < 0.05. (C) Mean fluorescence intensity (MFI) of platelet (i) GPVI and (ii) α2 integrin expression in CT (n=10), Shp1 KO (n=4), Shp2 KO (n=4) and Shp1/2 DKO (n=13) mice. Mean ± SEM; one way-ANOVA; *** P < 0.001.

The primary function of platelets is to prevent blood loss following injury. Therefore, to assess the impact of the loss of Shp1 and Shp2 on hemostasis, we performed tail bleeding assays. Although bleeding time was not significantly different between Shp1/2 DKO and control mice (Figure 2Bi), Shp1/2 DKO exhibited a significant increase in blood loss as reflected by higher hemoglobin levels (Figure 2Bii), indicating impaired hemostasis.

Aberrant functional responses of Shp1/2-deficient platelets

We next investigated platelet receptor expression by flow cytometry. The collagen activation receptor GPVI receptor expression was reduced in Shp1- and Shp2-deficient platelets compared with controls. However, the collagen activation receptor GPVI and the α2 subunit of the collagen integrin α2β1 were reduced by 25% and 13%, respectively, in Shp1/2-deficient platelets (Figure 2Ci-ii and Supplemental Table 3). All other major surface receptors analyzed were unaltered in these platelets (Supplemental Figure S2 and Supplemental Table 3).

We next checked platelet aggregation by light transmission aggregometry. Shp1- and Shp2-deficient platelets responded normally to all agonists tested (Figure 3A-C). However, consistent with reduced GPVI and α2 surface expression, Shp1/2-deficient platelets exhibited a marginal reduction in aggregation to a low and intermediate concentration of the GPVI-specific agonist CRP (1 and 3 μg/mL), and collagen (3 μg/mL), which signals via GPVI and binds with high affinity to the integrin α2β1 (Figure 3Ai-ii). However, they responded normally to thrombin (0.06 and 0.1 U/ml) (Figure 3Bi-ii), antibody-mediated cross-linking of the hem-ITAM-containing podoplanin receptor CLEC-2 (3 μg/mL) and ADP (3 μM) (Figure 3Ci-ii).

Aberrant platelet function in Shp1/2 DKO mice.

(A-C) Aggregation of washed platelets were measured by lumi-aggregometry in response to agonists indicated. Representative traces from independent experiments included in the quantitative analysis are shown. n= 4-8 mice/genotype per condition, percentage of aggregation at 5 minutes and area under the curve (AUC) quantification. Mean ± SEM, n = 5-17 mice/genotype per condition, one way-ANOVA, * P < 0.05. (D) Mean fluorescence intensity (MFI) of P-selectin expression of control (CT), Shp1 KO, Shp2 KO and Shp1/2 DKO platelets in whole blood in response to (i) 1 and 3 μg/mL CRP and (ii) 100 and 500 μM PAR4 peptide (PAR4p). Mean ± SEM, n = 5-9 mice/genotype per condition, two way-ANOVA; * P < 0.05, *** P < 0.001.

We also measured P-selectin surface expression in whole blood as a marker of α-granule secretion following platelet activation. Consistent with reduced GPVI expression, Shp1/2-deficient platelets showed markedly reduced P-selectin exposure in response to the GPVI-specific agonist CRP (1, 3, and 10 μg/mL) (Figure 3Di), whereas P-selectin expression in response to PAR-4 peptide (100 and 500 μM), which activates platelets through thrombin receptor signaling, was comparable to controls (Figure 3Dii).

Aberrant ITAM signaling in Shp1- and Shp2-deficient platelets

Since Shp1 and Shp2 have been implicated in regulating ITAM-containing receptor signaling, we investigated GPVI and CLEC-2 signaling in Shp1- and Shp2-deficient platelets. We focused on Src Family Kinases (SFK), which phosphorylate tyrosine residues within the ITAM of GPVI-associated FcR γ-chain and hemi-ITAM of the CLEC-2 receptor, and thereby enabling the recruitment and activation of Syk, which docks to both via its tandem SH2 domains, and mediates downstream effects. Activated Syk subsequently propagates the signal through phosphorylation of downstream effectors, including PLCγ2, leading to calcium mobilization and platelet activation. SFK activation was indirectly measured as trans-autophosphorylation of the highly conserved tyrosine residue 418 in Src (Src p-Tyr418) and Syk activation was measured as trans-autophosphorylation of tyrosine residues 519/520 (p-Tyr519/20), which directly correlate with activity. Western blot analysis demonstrated that Src p-Tyr418 was not altered under any conditions (Figure 4Ai-ii).. However, Syk p-Tyr519/520 was significantly decreased in CRP–stimulated platelets from Shp1/2-deficient platelets, and marginally reduced in CLEC-2-stimulated Shp1 KO and Shp2 KO deficient platelets, (Figure 4Bi-ii) suggesting that Shp1 and Shp2 modulate Syk-dependent signaling downstream of GPVI and CLEC-2.

Aberrant platelet signaling in Shp1/2 DKO mice.

Whole cell lysates of resting and (A) 10 µg/ml CRP-stimulated and (B) 10 µg/ml activating CLEC-2 antibody platelets from Shp1 KO, Shp2 KO and Shp1/2 DKO mice and litter-matched CT mice were western blotted with (A) anti-Src p-Ty418, Src total, (B) -Syk p-Tyr519/520 and Syk antibodies. (i) Representative blots of capillary-based immunoassays and (ii) quantification of peak areas from three independent experiments, Mean ± SEM, n = 3-4 mice/genotype; one-way ANOVA, ** P < 0.01, *** P < 0.001.

Impaired megakaryocyte differentiation and function

To determine the cause of thrombocytopenia in Shp1/2 DKO mice, we measured platelet recovery following antibody-mediated depletion as an indicator of platelet turnover. As expected, administration of the anti-GPIbα antibody resulted in extreme and sustained thrombocytopenia, confirming efficient depletion of circulating platelets (Supplemental Figure S3). No differences were observed the different phases or kinetics of platelet clearance or recovery in any of the mouse models. Interestingly, spleen size and weight were significantly increased in Shp1/2 DKO mice (Supplemental Figure S4), suggesting compensatory extramedullary hematopoiesis. However, HSPC colony-forming unit (CFU) assays performed on BM cells showed no significant difference in the frequency or composition of HSPCs, including MKPs between genotypes in vitro (Supplemental Figure S5).

Despite the maintenance of normal platelet turnover in vivo, in vitro analyses revealed pronounced defects in MK development, maturation and function of Shp1/2-deficient MKs (Figure 5A-C). BM-derived MKs from DKO mice exhibited a severe defect in polyploidization, as shown by a significant increase in 2-8N immature MKs, and decreased 16-64N mature MKs (Figure 5Ai-ii). Moreover, proplatelet formation was severely impaired, with a significantly lower percentage of MKs extending cytoplasmic projections compared to controls (Figure 5B).

Defects in megakaryocyte maturation and ex vivo proplatelet formation in Shp1/2 DKO.

(A) Mature BM-derived MKs from Shp1 KO, Shp2 KO, Shp1/2 DKO, and litter-matched CT mice were stained with propidium iodide and ploidy of cells was quantified by flow cytometry. (i-ii) The percentage of 2-4N and 8-128N ploidy cells was quantified (n = 4-6 mice/genotype; mean ± SEM; two-way ANOVA, * P < 0.05, ** P < 0.01, *** P < 0.001). (B) Ex vivo proplatelet formation. Percentage of MKs forming proplatelet was quantified in culture. Mean ± SEM, two-way ANOVA, *** P < 0.001. (C) Bone marrow explants. Proportion of MKs from Shp1 KO, Shp2 KO, Shp1/2 DKO and litter-matched CT mice extending proplatelets at 6h of observation were observed. Bars represent the mean ± SEM of six independent experiments. (i) Quantification and (ii) representatives’ images of round MKs, MKs with large ends and proplatelet (PPT) forming MKs. Scale bar: 50 µm. Mean ± SEM, one-way ANOVA, * P < 0.1, *** P < 0.001. (D) Classification of the MK according to their maturation stage: stage I (absence of granules), stage II (granules and developing demarcation membrane system (DMS) not yet organized), stage III (DMS organized in cytoplasmic territories). Data are reported as the percentage of the total number of MK, (i) of all stage MK and (ii) only stage 3 MK. Each bars segment represents the mean percentage ± SEM from three BM samples, *** P < 0.001. (E) Ploidy of in situ BM MKs measured by flow cytometry. Mean ± SEM, two-way ANOVA, * P < 0.1, *** P < 0.001.

Collectively, these data suggest that while homeostatic and compensatory thrombopoiesis is preserved in vivo, potentially supported by extramedullary hematopoiesis, Shp1 and Shp2 are required for efficient megakaryopoiesis in vitro, highlighting a critical role in late-stage megakaryopoiesis under defined conditions.

Defective proplatelet formation and MK maturation in Shp1/2-deficient mice

To further characterize the functional consequences of Shp1 and Shp2 deficiency on megakaryopoiesis, we evaluated proplatelet formation ex vivo and examined MK morphology in situ. We first performed ex vivo proplatelet formation assays using BM explants. The results revealed a marked reduction in proplatelet formation in Shp1/2 DKO mice compared with controls, with an increased number of round MKs and a decreased number of MK-forming proplatelets in Shp1/2 DKO mice (Figure 5Ci-ii). To assess MK maturation in situ, we performed EM and ploidy analysis in BM. In control mice, MK displayed characteristic features of terminal maturation, including large polyploid nuclei, demarcation membrane system (DMS) organization, and extensive cytoplasmic development. Shp1/2 DKO MK exhibited a maturation arrest, with increased frequency of stage III MK, defined by incomplete DMS formation, and limited cytoplasmic granularity (Figure 5Di-ii). This was further supported by flow cytometric analysis of DNA content, showing lower ploidy level (4N–8N) in Shp1/2 DKO MK (Figure 5E). These findings provide further evidence of the involvement of Shp1 and Shp2 in late-stage MK maturation and proplatelet formation, absence of which results in the accumulation of immature, stage III MK in the BM and a functional failure to complete thrombopoiesis.

Reduced Tpo signaling in Shp1/2-deficient MKs

To understand the molecular basis of impaired megakaryopoiesis in Shp1/2 DKO mice, we assessed Tpo-mediated signaling in BM-derived MKs. We previously demonstrated that Shp2 is a positive regulator of ERK1/2 downstream of Mpl. Tpo-mediated ERK1/2 activation was normal in Shp1-deficient MK and significantly reduced in Shp2- and Shp1/2-deficient MKs. However, other downstream effectors, including AKT and STAT3, showed modest or variable changes (Figure 6A-C). These findings provide a mechanistic explanation for the impaired MK development and maturation observed in vitro, and defective Tpo signaling through the ERK1/2 pathway in Shp1/2-deficient MK, underscoring the importance of these phosphatases in late-stage megakaryopoiesis.

Defects in TPO signaling in Shp1/2 DKO mice.

Mature BM-derived MKs from Shp1 KO, Shp2 KO, Shp1/2 DKO and litter-matched CT mice were stimulated with 50 ng/mL thrombopoietin (Tpo) for 10 min at 37°C. Whole cell lysates were western blotted with indicated antibodies A-C. Representative blots capillary-based immunoassays and quantification of peak areas from n = 3 independent experiments/genotype; two-way ANOVA, *** P < 0.001.

Shp2 allosteric inhibitors impair megakaryopoiesis and Tpo signaling

To further delineate the respective roles of Shp1 and Shp2 in MK and Tpo signaling, we employed two pharmacologically distinct inhibitors for each phosphatase to ensure specificity and validate findings from genetically modified mouse models.

For Shp1, we used M029, a covalent allosteric inhibitor that induces conformational changes that block Shp1 activity34 and F2Ac, a reversible inhibitor, which directly target the catalytic site (J.M. and Z.Y.Z., manuscript in preparation). Treatment of murine BM-derived MK with these Shp1 inhibitors showed no significant effects on cell viability, proliferation and MK ploidy (Figure 7Ai-iii), nor on Tpo-mediated signaling (Figure 7B), correlating with what was observed in Shp1-deficient MKs. Similarly, Shp2 was inhibited using SHP099 and RMC-4550, both selective allosteric inhibitors that stabilize Shp2 in its inactive conformation31. Addition of Shp2 inhibitors resulted in a marginal decrease in cell viability and a marked reduction of cell proliferation, both established functions of Shp2 in other lineages (Figure 7Ci-ii). Moreover, treatment with either SHP099 or RMC-4550 resulted in a significant inhibition of ploidy with a reduction in the percentage of cells achieving >8N (Figure 7Ciii), and a significant reduction in proplatelet formation (Figure 7Civ). This was accompanied by a severe impairment in Tpo-mediated ERK1/2 and AKT phosphorylation (Figure 7D). This dual-inhibitor profiling reveals that Shp2, but not Shp1, is critical for Tpo-driven MK proliferation, survival, endomitosis and Mpl signaling, highlighting Shp2 as a key regulator of late stage megakaryopoiesis and suggest that its inhibition disrupts thrombopoietic pathways.

Inhibition of Shp2 activity impairs megakaryopoiesis and Mpl signaling.

(A) Effects of two selective Shp1 allosteric inhibitors 10 µM M029 and 10 µM F2Ac were tested on megakaryopoeisis. (i) Viability, (ii) proliferation, and (iii) MK ploidy. Quantification from n = 3-4 independent experiments/condition. (B) Whole cell lysates of resting and 50 ng/mL Tpo-stimulated MKs were western blotted with the indicated antibodies. Representative blots capillary-based immunoassays and quantification of peak areas from n = 3 independent experiments/genotype. (C) Effects of two selective Shp2 allosteric inhibitors 10 µM SHP099 and 10 µM RMC-4550 were tested on megakaryopoeisis. (i) Viability, (ii) proliferation, (iii) MK ploidy and (iv) percentage of MKs forming proplatelets. Quantification from n = 3-6 independent experiments/condition. * P < 0.05; ** P < 0.01; *** P < 0.001, two-way ANOVA. (D) Whole cell lysates of resting and 50 ng/mL Tpo-stimulated MKs were western blotted with the indicated antibodies. Representative blots capillary-based immunoassays and quantification of peak areas from n = 3 independent experiments/genotype. * P < 0.05; ** P < 0.01; two-way ANOVA.

Discussion

In this study, we used the Gp1ba-Cre transgenic mouse model to achieve conditional deletion of the tyrosine phosphatases Shp1 and Shp2 in the MK/platelet lineage. Our results demonstrate efficient deletion of Shp1 and Shp2 in MKs and more completely in platelets, confirming the utility of this mouse model for studying protein involvement in megakaryopoiesis and thrombopoiesis. Residual Shp1 and Shp2 in MKs allowed us to study the phenotype of DKO MKs more thoroughly compared with DKO MKs on the Pf4-Cre background, which were severely compromised developmentally and did not survive in vitro.15 Additionally, the less severe MK defects meant fewer platelet anomalies, which contained lower levels of Shp1 and Shp2 than the MKs, suggesting different half-lives of the phosphatases in MKs versus platelets. The Gp1ba-Cre transgenic mouse allowed us to reveal distinct and synergistic functions of Shp1 and Shp2 in these processes, without confounding effects of deleting Shp1 and Shp2 in non-MK/platelet lineages observed with the Pf4-Cre transgene mouse.

Deletion of Shp1 or Shp2 alone did not significantly affect platelet counts or volumes, while combined deletion resulted in macrothrombocytopenia characterized by a 40% reduction in platelet counts and marginal increase in platelet volume. This phenotype was accompanied by prolonged bleeding and increased blood loss following tail tip excision, indicating impaired hemostatic function. These defects correlated with reduced GPVI and α2 expression, GPVI-mediated platelet aggregation and P-selectin expression. GPVI signaling was also impaired in Shp1/2-deficient platelets, with reduced Syk phosphorylation despite normal SFK activation, suggesting that Shp1 and Shp2 modulate ITAM/Syk proximal signaling (Figure 8). This is further supported by the marginal reduction in Syk activation downstream of the hemi-ITAM-containing CLEC-2 receptor. However, the marginal albeit significant reduction in Syk phosphorylation downstream of CLEC-2 in Shp1 and Shp2 KO platelets was not determined and was insufficient to impact CLEC-2-mediated platelet aggregation under the conditions tested. Differences in the stoichiometry and docking of Syk to phosphorylated GPVI-FcR γ-chain and CLEC-2 likely contribute to the differences in platelet reactivity and Syk phosphorylation downstream of the two receptors in the absence of Shp1 and Shp2.

Distinct roles of Shp1 and Shp2 in platelet GPVI signaling and megakaryocyte Tpo/Mpl signaling.

Schematic representation summarizing the effects of Shp1/Shp2 deficiency on platelet activation and megakaryopoiesis. In platelets, GPVI–FcRγ engagement by collagen induces ITAM phosphorylation and recruitment of Src family kinases and Syk, leading to platelet activation and aggregation under physiological conditions. Combined Shp1/Shp2 deficiency impairs GPVI expression and Syk activation, resulting in defective platelet activation and aggregation. In megakaryocytes, thrombopoietin (Tpo) binding to Mpl activates JAK2-dependent downstream signaling pathways, including STAT3/5, MAPK/ERK1/2, and PI3K/AKT, thereby promoting normal megakaryopoiesis and platelet production. Shp1/Shp2 deficiency or pharmacological inhibition of Shp2 reduces Tpo-induced ERK1/2 and AKT activation, leading to impaired megakaryocyte differentiation and thrombopoiesis.

Similar phases and kinetics of platelet recovery following immune-induced platelet clearance suggest that thrombocytopenia in Shp1/2 DKO mice is not due to increased platelet clearance, but rather to impaired MK development, maturation and function. In vitro assays showed significant defects in MK polyploidization and proplatelet formation, while in vivo analysis demonstrated an accumulation of immature MKs with incomplete demarcation membrane system formation and lower ploidy levels. These findings point to a critical requirement for Shp1 and Shp2 in late-stage MK maturation, a process tightly regulated by Tpo signaling through Mpl. Consistently, Tpo-mediated activation of the ERK1/2 pathway was markedly reduced in Shp2- and Shp1/2-deficient MKs, while changes in AKT and STAT3 pathways were modest, underscoring the prominent role of Shp2 in facilitating Tpo-driven Ras/MAPK signaling (Figure 8). Based on these findings, the synergistic effects of deleting both Shp1 and Shp2 suggests that Shp1 acts through another, as yet undefined pathway vital for megakaryopoiesis and thrombopoiesis. RhoA/ROCK is an obvious candidate that Shp1 has been implicated in regulating in other lineages and is essential for cytoskeletal remodeling during endomitosis.

Pharmacological inhibition corroborated genetic findings. Selective inhibition of Shp2 with allosteric inhibitors SHP099 and RMC-4550 significantly impaired MK viability, proliferation, polyploidization, proplatelet formation, and Tpo signaling, while Shp1 inhibitors M029 and F2Ac had no significant effects. This demonstrates a dominant role for Shp2 in regulating MK proliferation and differentiation downstream of Tpo, whereas Shp1 is dispensable for these functions. The pharmacological data thus confirm the functional specificity of the two phosphatases and provide further mechanistic insights into their differential contributions. The relatively mild platelet and hemostatic phenotypes observed in Shp1/2 DKO platelets suggest that Shp2 inhibition may have a limited impact on bleeding risk. However, given that these findings are derived from murine models, further studies are required to evaluate the hemostatic consequences of Shp2 inhibition in human systems and in disease-relevant contexts, which is part of a follow-up study.

The relatively milder, lineage-specific phenotype of Gp1ba-Cre-driven Shp1/2 DKO mice provides distinct advantages to the previously described Pf4-Cre mouse model 32,33 (Supplemental Table 4). Pf4-Cre-mediated deletion induces recombination earlier during megakaryopoiesis, starting in MK progenitors, whereas Gp1ba-Cre is activated predominantly in mature MKs and platelets. This temporal difference allows normal early MK development and platelet production, resulting in less severe thrombocytopenia and functional defects. Additionally, our protein quantification data show significant, but incomplete ablation of Shp1 and Shp2 in BM-derived MKs, suggesting residual phosphatase activity is sufficient under steady-state conditions. In contrast, Pf4-Cre models generally achieve near-complete deletion at earlier stages of development, leading to profound thrombocytopenia and often lethality. Furthermore, compensatory extramedullary hematopoiesis observed in our Gp1ba-Cre model may further alleviate thrombocytopenia. Importantly, this milder phenotype allowed us to study the Shp1/2 DKO mice in detail, which was impossible with the Pf4-Cre model due to the severity of the phenotype and cell viability. Findings further establish the utility of the Gp1ba-Cre strain for dissecting the distinct roles of Shp1 and Shp2 in late-stage megakaryopoiesis and platelet biology.

Collectively, Shp2 is a critical positive regulator of Mpl signaling and downstream Ras/MAPK pathways, essential for MK proliferation, polyploidization, and proplatelet formation, whereas Shp1 plays a more modulatory role in ITAM-containing receptor signaling in platelets without directly impacting MK maturation. The combined deletion of Shp1 and Shp2 reveals additive effects on platelet function and MK development, underscoring the importance of coordinated phosphatase regulation in maintaining platelet homeostasis. This study provides the first detailed analysis of Shp1 and Shp2 functions in megakaryopoiesis and platelet biology using a late-stage MK/platelet-specific deletion model, highlighting the indispensable role of Shp2 in megakaryopoiesis and thrombopoiesis, and distinct contributions of Shp1 in platelet signaling. It also emphasizes the utility of the Gp1ba-Cre mouse for exploring protein function in the MK/platelet lineage. These insights may have broader implications for understanding phosphatase regulation in hematopoiesis and could inform therapeutic targeting of Shp2 in hematologic diseases.